---
title: "3. Bin Visit Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Bin Visit Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(moo4feed)
library(ggplot2)
library(dplyr)
```

## 🚨 Important: Set Your Global Variables First!

*"Before we dive into exploring animal behavior," Ollie says while adjusting his tiny explorer hat, "make sure you've configured your global variables to match your data structure. This makes all the functions work smoothly!"*

If you haven't done this yet, run the setup from **Tutorial 1: Data Cleaning**:

```{r global-setup}
# 🎯 IMPORTANT: Set these to match YOUR data structure!
# Copy and modify these lines to match your column names and settings:

set_global_cols(
  id_col = "cow",           # Animal ID column name
  start_col = "start",      # Visit start time column  
  end_col = "end",          # Visit end time column
  bin_col = "bin",          # Bin/feeder ID column
  intake_col = "intake",    # Feed intake amount column
  dur_col = "duration",     # Visit duration column
  tz = "America/Vancouver"  # Your timezone
)

# Check that everything is set correctly:
cat("✅ Global variables set:\n")
cat("ID column:", id_col2(), "\n")
cat("Start time column:", start_col2(), "\n") 
cat("End time column:", end_col2(), "\n")
cat("Bin column:", bin_col2(), "\n")
cat("Intake column:", intake_col2(), "\n")
cat("Duration column:", duration_col2(), "\n")
cat("Timezone:", tz2(), "\n")
```

*"Perfect! Now all our functions will automatically use the right column names for your data!"* 🦦

## Ollie's Exploration Adventure 🦦

*"Hello again, my wonderful data explorers!"* Ollie exclaims, putting on his tiny safari hat. *"After solving the meal clustering mystery, I noticed something fascinating - some animals are adventurous explorers who love trying different feeding and drinking stations, while others are creatures of habit who stick to their favorite spots!"*

*"Think of it like this: imagine you're at a huge buffet. Some people (like me!) would visit every single station to try everything, while others might just stick to their favorite few dishes. Animals do the same thing with their feeding and drinking stations!"*

### What We'll Discover Today

- **Exploratory patterns**: Who are the adventurous animals vs. the creatures of habit?
- **Individual strategies**: How do different animals approach their environment?
- **Behavioral insights**: What can bin visit patterns tell us about animal preferences?

Let's dive into this exploration adventure! 🏊‍♂️

## Prerequisites: Clean Data First! 🧹

⚠️ **Important:** This tutorial assumes you have already completed:
- **Tutorial 1: Data Cleaning** 
- **Tutorial 2: Meal Clustering**

```{r data-prep}
# Load the example cleaned data from the package
data(clean_feed)
data(clean_water)

# If you're using your own data from previous tutorials, use this instead:
# clean_feed <- your_cleaned_feed_data     # From your cleaning results
# clean_water <- your_cleaned_water_data   # From your cleaning results

# Quick peek at our data structure
cat("Feed data structure:\n")
head(clean_feed[[1]], 3)  # First day, first 3 rows

cat("\nWater data structure:\n") 
head(clean_water[[1]], 3)  # First day, first 3 rows

cat("\nTotal days of data:\n")
cat("Feed days:", length(clean_feed), "\n")
cat("Water days:", length(clean_water), "\n")
```

## Understanding Exploratory Behavior 🔍

*"Here's where the magic happens!"* Ollie grins while polishing his magnifying glass. *"We're going to count how many different bins each animal visits. This tells us who the explorers are versus who the creatures of habit are!"*

Animals differ in how widely they explore their environment. Some visit many different feeding and drinking stations, while others prefer to stick to just a few favorite spots.

The `unique_bin_visits()` function helps us quantify this exploratory behavior:

```{r unique-bin-analysis}
# Calculate unique bin visits for each animal across all days
bin_visits <- unique_bin_visits(
  feed = clean_feed,            # Our cleaned feed data
  water = clean_water,          # Our cleaned water data
  return_list = FALSE)          # 🎯 Try changing to TRUE for day-by-day!

# Look at the results
head(bin_visits)
```

*"Isn't this exciting?"* Ollie claps his paws. *"Now we can see for each animal on each day:*
- *How many different **feed bins** they visited*
- *How many different **water bins** they visited*  
- *The **total number of unique bins** visited (feed + water)*

*This tells us which animals are 'explorers' versus 'creatures of habit'!"*

## Identifying the Most Exploratory Animals 🏃‍♀️

*"Now for the fun part - let's find our champion explorers!"* Ollie rubs his paws together excitedly.

Let's find the animals that consistently explore the most bins:

```{r exploratory-animals}
# Calculate the average number of unique bins visited per animal
avg_bin_visits <- bin_visits |>
  dplyr::group_by(cow) |>
  dplyr::summarize(
    avg_feed_bins = mean(unique_feed_bins_visited),
    avg_water_bins = mean(unique_water_bins_visited),
    avg_total_bins = mean(total_bins_visited),
    days_observed = dplyr::n()
  ) |>
  dplyr::arrange(dplyr::desc(avg_total_bins))

# Display the top 10 most exploratory animals
cat("🏆 TOP 10 MOST EXPLORATORY ANIMALS:\n")
head(avg_bin_visits, 10)

cat("\n🏠 TOP 10 CREATURES OF HABIT (fewest bins visited):\n")
tail(avg_bin_visits, 10)
```

*"Look at that!"* Ollie points excitedly at the results. *"We can clearly see the difference between our adventurous explorers and our creatures of habit. Each animal has their own unique strategy for navigating their environment!"*

## 🎉 Ollie's Exploration Insights!

*"Wow! What an adventure!"* Ollie exclaims, looking at all the exploration data. *"We've discovered so much about animal behavior patterns! Now we can clearly see:*

- *Who the champion explorers are (they visit lots of different bins!)*
- *Who the creatures of habit are (they stick to their favorites)*  
- *Individual animal personalities and strategies*

*"This information is super valuable for understanding animal welfare, management decisions, and individual needs!"*

**What we've accomplished together:**
- ✅ Learned to quantify exploratory behavior
- ✅ Identified individual animal exploration strategies  
- ✅ Discovered patterns in animal preferences

*"The best part is you can now use these techniques to understand your own animals better!"* 🦦

## 🦦 Bin Visit Analysis Cheat Sheet

*"Here's my handy reference guide for bin visit analysis! Copy and paste these code blocks to work with your own data."*

```r
#' ============================================================================
#' 🦦 OLLIE'S BIN VISIT ANALYSIS CHEAT SHEET
#' ============================================================================
#' Copy and modify these code blocks for your own analysis!

# ---- SETUP: Global Variables (REQUIRED FIRST!) ----
library(moo4feed)
library(dplyr)

# Set up your column names and timezone (modify these!)
set_global_cols(
  id_col = "cow",           # Your animal ID column
  start_col = "start",      # Visit start time column  
  end_col = "end",          # Visit end time column
  bin_col = "bin",          # Bin/feeder ID column
  intake_col = "intake",    # Feed intake amount column
  dur_col = "duration",     # Visit duration column
  tz = "America/Vancouver"  # Your timezone
)

# ---- STEP 1: Load Your Data ----
# Use your own cleaned data from previous tutorials:
# clean_feed <- your_cleaned_feed_data
# clean_water <- your_cleaned_water_data

# ---- STEP 2: Calculate Unique Bin Visits ----
# Get overall summary across all days
bin_visits <- unique_bin_visits(
  feed = your_clean_feed,           # Your cleaned feed data
  water = your_clean_water,         # Your cleaned water data  
  return_list = FALSE               # FALSE = combined summary, TRUE = day-by-day
)

# ---- STEP 3: Identify Most/Least Exploratory Animals ----
# Calculate averages per animal
avg_bin_visits <- bin_visits |>
  dplyr::group_by(cow) |>
  dplyr::summarize(
    avg_feed_bins = mean(unique_feed_bins_visited),
    avg_water_bins = mean(unique_water_bins_visited), 
    avg_total_bins = mean(total_bins_visited),
    days_observed = dplyr::n(),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(avg_total_bins))    # Most exploratory first

# Get top and bottom explorers
top_explorers <- head(avg_bin_visits, 10)       # Most exploratory
creatures_of_habit <- tail(avg_bin_visits, 10)  # Least exploratory

# ---- STEP 4: Quick Analysis ----
# Check exploration results
head(avg_bin_visits, 10)               # Top 10 explorers
tail(avg_bin_visits, 10)               # Top 10 creatures of habit

#' ============================================================================
#' 🦦 That's it! You now understand animal exploration patterns!
#' ============================================================================
``` 