---
title: "3. Bin Visit Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Bin Visit Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(moo4feed)
library(ggplot2)
library(dplyr)
```

## 1. 🚨 Important: Set Your Global Variables First!

Before analyzing bin visit patterns, configure global variables to match your data structure:

```{r global-setup}
# Configure global variables for your data structure
set_global_cols(
  id_col = "cow",           # Animal ID column name
  start_col = "start",      # Visit start time column  
  end_col = "end",          # Visit end time column
  bin_col = "bin",          # Bin/feeder ID column
  intake_col = "intake",    # Feed intake amount column
  dur_col = "duration",     # Visit duration column
  tz = "America/Vancouver"  # Your timezone
)

# Verify configuration
cat("✅ Global variables configured:\n")
cat("ID column:", id_col2(), "\n")
cat("Start time column:", start_col2(), "\n") 
cat("End time column:", end_col2(), "\n")
cat("Bin column:", bin_col2(), "\n")
cat("Intake column:", intake_col2(), "\n")
cat("Duration column:", duration_col2(), "\n")
cat("Timezone:", tz2(), "\n")
```

## 2. Introduction to Bin Visit Analysis

Ollie the Otter, our behavioral analysis specialist, explains: "Animals exhibit distinct exploration patterns when accessing feeding and drinking stations. Some individuals visit many different bins while others consistently use only a few preferred locations. This behavior reveals individual personalities and resource utilization strategies."

## 3. Prerequisites

This tutorial assumes completion of previous data processing steps:
- **Tutorial 1: Data Cleaning** 
- **Tutorial 2: Meal Clustering**

## 4. Data Preparation

```{r data-prep}
# Load cleaned example data
data(clean_feed)
data(clean_water)

# If you're using your own data from previous tutorials, use this instead:
# clean_feed <- your_cleaned_feed_data     # From your cleaning results
# clean_water <- your_cleaned_water_data   # From your cleaning results

# Quick peek at our data structure
cat("Feed data structure:\n")
head(clean_feed[[1]], 3)  # First day, first 3 rows

cat("\nWater data structure:\n") 
head(clean_water[[1]], 3)  # First day, first 3 rows

cat("\nTotal days of data:\n")
cat("Feed days:", length(clean_feed), "\n")
cat("Water days:", length(clean_water), "\n")
```

## 5. Exploratory Behavior Analysis

Individual animals differ in their willingness to explore different feeding and drinking stations. This analysis quantifies exploration patterns by counting unique bin visits per animal per day.

### Calculate Unique Bin Visits

```{r unique-bin-analysis}
# Calculate unique bin visits for each animal across all days
bin_visits <- unique_bin_visits(
  feed = clean_feed,            # Our cleaned feed data
  water = clean_water,          # Our cleaned water data
  return_list = FALSE)          # 🎯 Try changing to TRUE for day-by-day!

# Look at the results
head(bin_visits)
```

This analysis provides three key metrics for each animal on each day:
- *How many different **feed bins** they visited*
- *How many different **water bins** they visited*  
- *The **total number of unique bins** visited (feed + water)*

## 6. Identifying Exploration Patterns

### Most Exploratory Animals

```{r exploratory-animals}
# Calculate average unique bin visits per animal
avg_bin_visits <- bin_visits |>
  dplyr::group_by(cow) |>
  dplyr::summarize(
    avg_feed_bins = mean(unique_feed_bins_visited),
    avg_water_bins = mean(unique_water_bins_visited),
    avg_total_bins = mean(total_bins_visited),
    days_observed = dplyr::n(),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(avg_total_bins))

# Display most exploratory animals
cat("🏆 TOP 10 MOST EXPLORATORY ANIMALS:\n")
head(avg_bin_visits, 10)

cat("\n🏠 TOP 10 CREATURES OF HABIT (fewest bins visited):\n")
tail(avg_bin_visits, 10)
```

This analysis reveals clear individual differences in exploration behavior. Most exploratory animals visit many different stations, while creatures of habit consistently use fewer preferred locations.


## 7. Summary

This tutorial demonstrated bin visit analysis for understanding animal exploration behavior:

✅ **Quantified exploration**: Measured individual differences in station usage
✅ **Identified patterns**: Distinguished exploratory vs. habitual behaviors  
✅ **Research applications**: Connected patterns to welfare and management implications

The analysis provides valuable insights for optimizing facility design, monitoring animal welfare, and understanding individual behavioral differences.

## 8. Code Reference

```r
#' Copy and modify these code blocks for your own analysis!

# ---- SETUP: Global Variables (REQUIRED FIRST!) ----
library(moo4feed)
library(dplyr)

# Set up your column names and timezone (modify these!)
set_global_cols(
  id_col = "cow",           # Your animal ID column
  start_col = "start",      # Visit start time column  
  end_col = "end",          # Visit end time column
  bin_col = "bin",          # Bin/feeder ID column
  intake_col = "intake",    # Feed intake amount column
  dur_col = "duration",     # Visit duration column
  tz = "America/Vancouver"  # Your timezone
)

# ---- STEP 1: Load Your Data ----
# Use your own cleaned data from previous tutorials:
# clean_feed <- your_cleaned_feed_data
# clean_water <- your_cleaned_water_data

# ---- STEP 2: Calculate Unique Bin Visits ----
# Get overall summary across all days
bin_visits <- unique_bin_visits(
  feed = your_clean_feed,           # Your cleaned feed data
  water = your_clean_water,         # Your cleaned water data  
  return_list = FALSE               # FALSE = combined summary, TRUE = day-by-day
)

# ---- STEP 3: Identify Most/Least Exploratory Animals ----
# Calculate averages per animal
avg_bin_visits <- bin_visits |>
  dplyr::group_by(cow) |>
  dplyr::summarize(
    avg_feed_bins = mean(unique_feed_bins_visited),
    avg_water_bins = mean(unique_water_bins_visited), 
    avg_total_bins = mean(total_bins_visited),
    days_observed = dplyr::n(),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(avg_total_bins))    # Most exploratory first

# Get top and bottom explorers
top_explorers <- head(avg_bin_visits, 10)       # Most exploratory
creatures_of_habit <- tail(avg_bin_visits, 10)  # Least exploratory

# ---- STEP 4: Quick Analysis ----
# Check exploration results
head(avg_bin_visits, 10)               # Top 10 explorers
tail(avg_bin_visits, 10)               # Top 10 creatures of habit

``` 