---
title: "8. Meal-Level Behavior Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{8. Meal-Level Behavior Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(moo4feed)
library(ggplot2)
library(dplyr)
```

## 1. Set Your Global Variables First

Before analyzing meal-level behavior patterns, configure global variables to match your data structure:

```{r global-setup}
# Configure global variables for your data structure
set_global_cols(
  id_col = "cow",           # Animal ID column name
  start_col = "start",      # Visit start time column
  end_col = "end",          # Visit end time column
  bin_col = "bin",          # Bin/feeder ID column
  intake_col = "intake",    # Feed intake amount column
  dur_col = "duration",     # Visit duration column
  tz = "America/Vancouver"  # Your timezone
)

```

## 2. Introduction to Meal-Level Behavior Analysis

A meal groups multiple feeding visits that happened consecutively in time. Analyzing feeding behaviours on meal-level instaed of visit-level is more biologically meaningful. This tutorial demonstrates how to:

1. **Analyze non-nutritive visits within meals** - Identify exploratory behavior during each meal
2. **Examine actor/reactor roles within meals** - Understand the frequency of each animal being an actor/reactor for agonistic interactions during meals
3. **Summarize meal-level behavior patterns** - Aggregate meal-level behavior patterns per animal per day

## 3. Prerequisites

This tutorial requires:

1. Completion of [**Tutorial 1: Data Cleaning**](data_cleaning.html)
2. Completion of [**Tutorial 2: Meal Clustering**](meal_clustering.html) - your data must have meal labels
3. For actor/reactor analysis: [**Tutorial 4: Replacement Detection**](replacement_detection.html)

## 4. Data Preparation

First, we need to prepare meal-labeled data:

```{r data-prep}
# Load cleaned example data
data(clean_feed)
data(clean_comb)

# If you're using your own data from previous tutorials, use this instead:
# clean_feed <- your_cleaned_feed_data
# clean_comb <- your_cleaned_comb_data

# Label visits with meal assignments
# (See Tutorial 2: Meal Clustering for details on parameter selection)
labeled_visits <- meal_label_visits(
  data = clean_feed,
  eps = NULL,                # Auto-determine optimal interval
  min_pts = 2,               # Minimum visits to form a meal
  method = "gmm",            # Use GMM for interval detection
  eps_scope = "all_animals",
  lower_bound = NULL,
  upper_bound = NULL,
  use_log_transform = TRUE,
  log_multiplier = 20,
  log_offset = 1
)

# Quick peek at meal-labeled data
labeled_visits[[1]][, c("cow", "bin", "start", "intake", "meal_id", "meal_start")] |>
  dplyr::arrange(cow, meal_id, start) |>
  head()

```

## 5. Non-Nutritive Visits Within Meals

Non-nutritive visits (visits where animals don't consume feed despite it being available) and empty bin visits (visits to bins with no feed) can occur during meals. Analyzing these patterns within meal contexts provides insights into exploratory behavior.

### Understanding Visit Types Within Meals

The `meal_non_nutritive_summary()` function classifies each visit within a meal as:

- **Nutritive**: Animal consumed feed (intake > calibration error)
- **Non-nutritive**: Feed available but animal didn't eat (exploratory)
- **Empty bin**: No feed available in the bin

### Analyze Non-Nutritive Patterns

```{r non-nutritive-analysis}
# Create quality control configuration
my_qc_config <- qc_config(
  calibration_error = 0.5    # Equipment measurement threshold (kg)
)

# Analyze non-nutritive and empty bin visits within meals
meal_visits <- meal_non_nutritive_summary(
  data = labeled_visits,
  cfg = my_qc_config
)

# Examine the results from first day, sort by total non-nutritive visits descending
meal_visits[[1]] |>
  dplyr::arrange(desc(total_non_nutritive_visits)) |>
  head()
```

### Visualize Non-Nutritive Patterns

```{r viz-non-nutritive-bar, fig.width=10, fig.height=6}
# Use first day's data as an example
first_day_data <- meal_visits[[1]] |>
  dplyr::arrange(desc(total_non_nutritive_visits)) |>
  head(50)  # Limit to top 50 animals

# Create bar plot
ggplot(first_day_data, aes(x = reorder(cow, total_non_nutritive_visits), 
                           y = mean_non_nutritive_per_meal)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  coord_flip() +  # Horizontal bars for better readability
  labs(
    title = "Average Non-Nutritive Visits Per Meal by Animal",
    subtitle = "Top 50 animals sorted by total non-nutritive visits (first day example)",
    x = "Animal ID",
    y = "Average Non-Nutritive Visits Per Meal"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11)
  )
```

The summary provides per animal per day:

- **`mean_non_nutritive_per_meal`**: Average non-nutritive visits per meal
- **`median_non_nutritive_per_meal`**: Median non-nutritive visits per meal
- **`sd_non_nutritive_per_meal`**: Standard deviation
- **`mean_empty_bin_per_meal`**: Average empty bin visits per meal
- **`median_empty_bin_per_meal`**: Median empty bin visits per meal
- **`sd_empty_bin_per_meal`**: Standard deviation
- **`total_non_nutritive_visits`**: Total count of non-nutritive visits
- **`total_empty_bin_visits`**: Total count of empty bin visits
- **`total_meals`**: Number of meals analyzed

### Visualize Non-Nutritive Patterns

```{r viz-non-nutritive, fig.width=6, fig.height=4}
# Combine all days all animals
all_meal_visits <- do.call(rbind, meal_visits)

# Distribution of non-nutritive visits per meal across all animals
ggplot(all_meal_visits, aes(x = mean_non_nutritive_per_meal)) +
  geom_histogram(bins = 20, fill = "coral", alpha = 0.7) +
  labs(
    title = "Distribution of Non-Nutritive Visits Per Meal",
    x = "Average Non-Nutritive Visits Per Meal",
    y = "Frequency"
  ) +
  theme_minimal()
```

## 6. Actor/Reactor Roles Within Meals

When we combine meal data with replacement events, we can analyze the social roles animals play **during their meals**. An animal might enter a visit as an "actor" (replacing another animal) or leave as a "reactor" (being replaced by another animal).

### Understanding Actor/Reactor Roles

- **Actor**: The animal entered this visit by replacing another animal at the bin
- **Reactor**: The animal was replaced by another animal when leaving this visit
- **Actor-Reactor**: Both events occurred during the same visit (rare but possible). They entered as actor and left as reactor within the same visit.

### Prepare Replacement Data

```{r replacement-prep}
# Detect replacement events
# (See Tutorial 4: Replacement Detection for details)
replacements <- record_replacement_days(
  comb = clean_comb,
  cfg = qc_config(replacement_threshold = 26)
)

# Replacement events per day:
sapply(replacements, nrow)
```

### Analyze Actor/Reactor Roles Within Meals

We need meal-labeled data that matches the combined feed/water data used for replacement detection:

```{r role-analysis}
# Label the combined data with meals
labeled_comb <- meal_label_visits(
  data = clean_comb,
  eps = NULL,
  min_pts = 2,
  method = "gmm",
  eps_scope = "all_animals",
  lower_bound = NULL,
  upper_bound = NULL,
  use_log_transform = TRUE,
  log_multiplier = 20,
  log_offset = 1
)

# Analyze actor/reactor roles within meals
meal_roles <- meal_replacement_roles(
  visit_data = labeled_comb,
  replacement_data = replacements,
  time_tolerance = 1           # Seconds tolerance for matching
)

# Examine the meal-level results
# Actor/reactor roles per meal (first day)
meal_roles[[1]] |>
  dplyr::arrange(desc(pct_reactor)) |>
  head()
```

The results show for each animal's meal:

- **`total_visits_in_meal`**: Total feeding visits in that meal
- **`actor_visits`**: Total number of visits where animal entered as actor in that meal
- **`reactor_visits`**: Total number of visits where animal left as reactor in that meal
- **`actor_reactor_visits`**: Total number of visits when they entered as actor and left as reactor in that meal
- **`pct_actor`**: Percentage of visits as actor in that meal
- **`pct_reactor`**: Percentage of visits as reactor in that meal
- **`pct_actor_reactor`**: Percentage of visits when they entered as actor and left as reactor in that meal

### Summarize Roles Per Animal Per Day

```{r role-summary}
# Get daily summary statistics across all meals
daily_roles <- meal_replacement_roles_summary(meal_roles)

# Daily role summary (first day)
daily_roles[[1]] |>
  dplyr::arrange(desc(mean_pct_actor)) |>
  head()
```

The daily summary provides:

- **`date`**: Date of the summary
- **`cow`**: Animal identifier
- **`mean_pct_actor`**: Average percentage of visits as an actor across all meals
- **`median_pct_actor`**: Median percentage of visits as an actor across all meals
- **`sd_pct_actor`**: Standard deviation of percentage of visits as an actor across all meals
- **`mean_pct_reactor`**: Average percentage of visits as a reactor across all meals
- **`median_pct_reactor`**: Median percentage of visits as a reactor across all meals
- **`sd_pct_reactor`**: Standard deviation of percentage of visits as a reactor across all meals
- **`mean_pct_actor_reactor`**: Average percentage of visits with both actor and reactor roles across all meals
- **`median_pct_actor_reactor`**: Median percentage of visits with both actor and reactor roles across all meals
- **`sd_pct_actor_reactor`**: Standard deviation of percentage of visits with both actor and reactor roles across all meals
- **`total_meals`**: Number of meals analyzed


## 8. Summary

This tutorial demonstrated meal-level behavior analysis:

- **Non-nutritive visits within meals**: Identified exploratory behavior patterns
- **Actor/reactor roles**: Analyzed the percentage of visits as an actor or reactor across all meals
- **Daily summaries**: Aggregated behavior patterns per animal per day

## 9. Code Cheatsheet

```{r cheatsheet, eval=FALSE}
#' Copy and modify these code blocks for your own analysis!

# ---- SETUP: Global Variables (REQUIRED FIRST!) ----
library(moo4feed)
library(ggplot2)
library(dplyr)

# Set up your column names and timezone (modify these!)
set_global_cols(
  id_col = "cow",           # Your animal ID column
  start_col = "start",      # Visit start time column
  end_col = "end",          # Visit end time column
  bin_col = "bin",          # Bin/feeder ID column
  intake_col = "intake",    # Feed intake amount column
  dur_col = "duration",     # Visit duration column
  tz = "America/Vancouver"  # Your timezone
)

# ---- STEP 1: Prepare Meal-Labeled Data ----
# Load your cleaned data
data(clean_feed)
data(clean_comb)

# Label visits with meals (see Tutorial 2 for details)
labeled_visits <- meal_label_visits(
  data = clean_feed,
  eps = NULL,                # Auto-determine
  min_pts = 2,
  method = "gmm",
  eps_scope = "all_animals",
  use_log_transform = TRUE,
  log_multiplier = 20,
  log_offset = 1
)

# ---- STEP 2: Analyze Non-Nutritive Visits Within Meals ----
my_qc_config <- qc_config(calibration_error = 0.5)

meal_visits <- meal_non_nutritive_summary(
  data = labeled_visits,
  cfg = my_qc_config
)

# View results
head(meal_visits[[1]])

# Combine all days
all_meal_visits <- do.call(rbind, meal_visits)

# Find animals with high exploratory behavior
high_exploratory <- all_meal_visits |>
  dplyr::group_by(cow) |>
  dplyr::summarise(
    avg_non_nutritive = mean(mean_non_nutritive_per_meal, na.rm = TRUE),
    avg_empty_bin = mean(mean_empty_bin_per_meal, na.rm = TRUE),
    .groups = "drop"
  ) |>
  dplyr::arrange(desc(avg_non_nutritive))

print(high_exploratory)

# ---- STEP 3: Analyze Actor/Reactor Roles Within Meals ----
# First, detect replacement events (see Tutorial 4)
replacements <- record_replacement_days(
  comb = clean_comb,
  cfg = qc_config(replacement_threshold = 26)
)

# Label combined data with meals
labeled_comb <- meal_label_visits(
  data = clean_comb,
  eps = NULL,
  min_pts = 2,
  method = "gmm",
  eps_scope = "all_animals"
)

# Analyze roles within meals
meal_roles <- meal_replacement_roles(
  visit_data = labeled_comb,
  replacement_data = replacements,
  time_tolerance = 1
)

# View meal-level results
head(meal_roles[[1]])

# ---- STEP 4: Get Daily Role Summaries ----
daily_roles <- meal_replacement_roles_summary(meal_roles)

# View daily summaries
print(daily_roles[[1]])

# Combine all days
all_daily_roles <- do.call(rbind, daily_roles)

# Find dominant/subordinate animals
dominance_summary <- all_daily_roles |>
  dplyr::group_by(cow) |>
  dplyr::summarise(
    avg_pct_actor = mean(mean_pct_actor, na.rm = TRUE),
    avg_pct_reactor = mean(mean_pct_reactor, na.rm = TRUE),
    total_meals = sum(total_meals),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    dominance_ratio = avg_pct_actor / (avg_pct_reactor + 0.01)
  ) |>
  dplyr::arrange(desc(dominance_ratio))

print(dominance_summary)

# ---- STEP 5: Visualize Results ----
# Distribution of non-nutritive visits
ggplot(all_meal_visits, aes(x = mean_non_nutritive_per_meal)) +
  geom_histogram(bins = 20, fill = "coral", alpha = 0.7) +
  theme_minimal()

# Actor vs Reactor scatter plot
ggplot(dominance_summary, aes(x = avg_pct_reactor, y = avg_pct_actor)) +
  geom_point(aes(size = total_meals), alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  theme_minimal()
```

---
