---
title: "3. Meal Clustering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Meal Clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(moo4feed)
library(ggplot2)
library(dplyr)
library(patchwork)
```

## Ollie's Meal Clustering Journey 🦦

*Our favorite data otter Ollie is back! This time, he's puzzled by something he noticed while watching the cows at the research farm...*

"I've been tracking when each cow visits the feeders," Ollie says, adjusting his tiny data analyst glasses, "but I realized something important - just because a cow makes two separate visits doesn't mean they're having two separate meals! Sometimes they just take a quick break and come back."

"So I thought, what if we could group visits into meals and visualize each cow's breakfast, lunch and dinner?"

### The Meal Clustering Challenge

Individual feeding visits tell only part of the story. To understand animal feeding behavior, we need to group these visits into meaningful **meals** - periods of sustained feeding activity separated by longer breaks.

Ollie's goal is to help you:
1. **Analyze temporal patterns** - Look at time gaps between consecutive visits
2. **Find natural breakpoints** - Identify optimal time thresholds that separate within-meal from between-meal visit gaps  
3. **Group visits intelligently** - Use clustering algorithms to group nearby visits into meals
4. **Summarize behavior** - Create meal-level summaries for analysis

*"Let's solve this puzzle together!"* 🦦

## Understanding Inter-Visit Gaps

*"First," says Ollie, "we need to understand the pattern of time gaps between feeding visits. Are there short gaps (within meals) and long gaps (between meals)?"*

Let's help Ollie examine some example data:

```{r data-prep}
# Use a subset of animals for demonstration
demo_data <- clean_feed

# Quick look at our data
cat("Cows on the third day:", unique(demo_data[[3]]$cow), "\n")
cat("Dates with data:", names(demo_data), "\n")
```

## Method 1: Percentile-Based Interval Selection

The simplest approach uses percentiles of inter-visit gaps to determine the optimal meal interval (eps).

### How It Works

The percentile method:
1. Calculates gaps between consecutive visits for each animal within each day
2. Takes a high percentile (default 93rd) as the threshold
3. Visits separated by less than this threshold belong to the same meal

```{r percentile-method}
# Calculate optimal interval using percentile method
eps_percentile <- meal_interval(demo_data, 
                               method = "percentile", 
                               percentile = 0.93,
                               id_col = "cow", 
                               start_col = "start", 
                               end_col = "end",
                               tz = "America/Vancouver")

cat("Optimal interval (93rd percentile):", round(eps_percentile, 2), "minutes\n")
```

### Visualizing the Percentile Method

```{r viz-percentile, fig.width=10, fig.height=6}
# Visualize gap distribution with percentile-based eps
# 🎯 TRY THIS: Change these percentile values!
# Try changeing the percentile to 0.75, 0.85, 0.95, 0.99 and see how it affects the threshold!
p1 <- viz_eps_percentile(demo_data, 
                        percentile = 0.93,
                        id_col = "cow", 
                        start_col = "start", 
                        end_col = "end",
                        tz = "America/Vancouver",
                        xlim = 200)

p1
```

## Method 2: Gaussian Mixture Model (GMM) Approach

*"Wait!" Ollie's eyes light up. "I just learned about a fancier statistical method! What if we fit mathematical curves to our gap data to find where short gaps and long gaps naturally separate?"*

### How Ollie's GMM Method Works

The GMM method is like fitting two bell curves to our data:
1. Fits a 2-component Gaussian mixture model to inter-visit gaps
2. Identifies two populations: **within-meal gaps** (short) and **between-meal gaps** (long)
3. Finds the intersection point between these distributions as the optimal threshold
4. Can use log-transformation to better handle skewed gap distributions

*"It's like finding where two waves cross!"* 🌊

```{r gmm-method}
# Calculate optimal interval using GMM method
eps_gmm <- meal_interval(demo_data, 
                        method = "gmm", 
                        use_log_transform = TRUE,
                        id_col = "cow", 
                        start_col = "start", 
                        end_col = "end",
                        tz = "America/Vancouver")

cat("Optimal interval (GMM):", round(eps_gmm, 2), "minutes\n")
```

### Visualizing the GMM Method

```{r viz-gmm, fig.width=10, fig.height=6}
# Visualize gap distribution with GMM fit
p2 <- viz_eps_gmm(demo_data, 
                  use_log_transform = TRUE,
                  id_col = "cow", 
                  start_col = "start", 
                  end_col = "end",
                  tz = "America/Vancouver",
                  xlim = 10)  # Using smaller xlim for log-transformed data

p2
```


### 🎯 Experiment with Log Transformations in GMM!

*"The log transformation has some knobs we can twist!" Ollie says excitedly. "Let's see how different parameters affect our results!"*

**🔄 Interactive Challenge**: Try different log transformation parameters:

```{r gmm-parameters}
# 🎯 TRY THIS: Change these parameter combinations!
# Try different multipliers: 5, 10, 20, 30, 50
# Try different offsets: 0.5, 1, 2
log_params <- list(
  list(multiplier = 10, offset = 1),    # 🎯 Change these values!
  list(multiplier = 20, offset = 1),    # 🎯 Change these values!
  list(multiplier = 50, offset = 1)     # 🎯 Change these values!
)

for (i in seq_along(log_params)) {
  eps_val <- meal_interval(demo_data, 
                          method = "gmm", 
                          use_log_transform = TRUE,
                          log_multiplier = log_params[[i]]$multiplier,  # 🎯 Your input here!
                          log_offset = log_params[[i]]$offset,         # 🎯 Your input here!
                          id_col = "cow", 
                          start_col = "start", 
                          end_col = "end",
                          tz = "America/Vancouver")
  
  cat("Log params (mult=", log_params[[i]]$multiplier, ", offset=", log_params[[i]]$offset, 
      "): eps =", round(eps_val, 2), "minutes\n")
}
```

### Visualize different Log Transformations

Log transformation often improves GMM fitting for highly skewed gap distributions:

```{r gmm-comparison, fig.width=12, fig.height=8}
# Compare GMM with and without log transformation
p_gmm_no_log <- viz_eps_gmm(demo_data, 
                           use_log_transform = FALSE,
                           id_col = "cow", 
                           start_col = "start", 
                           end_col = "end",
                           tz = "America/Vancouver",
                           xlim = 200)

# with basic log transformation: log(x + 1) where x is the gap in minutes
p_gmm_log <- viz_eps_gmm(demo_data, 
                        use_log_transform = TRUE,
                        log_multiplier = 1,
                        log_offset = 1,
                        id_col = "cow", 
                        start_col = "start", 
                        end_col = "end",
                        tz = "America/Vancouver",
                        xlim = 8)

# with customzied log transformation: log(20x + 1) where x is the gap in minutes
p_gmm_log_20 <- viz_eps_gmm(demo_data, 
                        use_log_transform = TRUE,
                        log_multiplier = 20,
                        log_offset = 1,
                        id_col = "cow", 
                        start_col = "start", 
                        end_col = "end",
                        tz = "America/Vancouver",
                        xlim = 8)

# Combine plots
patchwork::wrap_plots(p_gmm_no_log, p_gmm_log, p_gmm_log_20, ncol = 1)
```

**💡 Ollie's Tip**: *"Different parameters can give different results. The key is finding what works best for your specific data!"*

## DBSCAN: Ollie's Clustering Algorithm

*"Now comes the exciting part!" Ollie grins. "Once we know our optimal interval, we need to actually group the visits into meals. I'll use something called DBSCAN - it's like finding dense neighborhoods of visits!"*

### How Ollie's DBSCAN Works

DBSCAN is like Ollie looking for "busy neighborhoods of visits":

1. **Core points**: Visits with enough nearby friends (at least `min_pts` neighbors within `eps` time)
2. **Border points**: Visits near a busy neighborhood but not busy themselves
3. **Noise points**: Lonely visits that don't belong to any meal (isolated visits)
4. **Clusters**: Groups of connected busy neighborhoods = **MEALS!**

For meal clustering:
- **Points** = feeding visit start times 🕐
- **Distance** = time difference between the end time of the previous visit and the start time of the current visit ⏱️
- **eps** = maximum time gap within a meal ⏰
- **min_pts** = minimum visits to form a dense meal cluster 🍽️

*"It's like connecting the dots, but only if the dots are close enough together!"* 🦦

## Clustering Visits into Meals

Now let's cluster our feeding visits into meals using different parameters:

### Basic Clustering

```{r basic-clustering}
# Cluster meals with default parameters
# 🎯 TRY THIS: Change these min_pts values!
# Try: c(1, 2, 3, 4) or c(2, 3, 5, 7) or even c(1, 2, 4, 8)
# Rule of thumb: min_pts >= D + 1 where D is the number of dimensions of the data.
# So for us min_pts = 1 + 1 = 2
meals_basic <- cluster_meals(demo_data,
                            eps = NULL,  # Auto-determine using GMM
                            min_pts = 2,
                            id_col = "cow",
                            start_col = "start",
                            end_col = "end",
                            bin_col = "bin",
                            intake_col = "intake",
                            dur_col = "duration",
                            tz = "America/Vancouver")

# Look at the results
head(meals_basic)
cat("Total meals identified:", nrow(meals_basic), " in 3 days\n")
cat("Meals per animal:\n")
head(table(meals_basic$cow, meals_basic$date))
```


### Different eps Scopes

You can calculate optimal eps using different data scopes:
- **all_animals** = calculate a universal optimal eps using visit gaps for all animals from all days
- **one_animal_all_days** = Customize eps for each individual animal, calculate eps based on every animal across all days
- **one_animal_single_day** = Customize eps for each individual and different days, calculate eps based on each animal on each day

```{r eps-scopes}
# Compare different eps calculation scopes
scopes <- c("all_animals", "one_animal_all_days", "one_animal_single_day")
scope_results <- list()

for (scope in scopes) {
  meals_temp <- cluster_meals(demo_data,
                             eps = NULL,
                             eps_scope = scope,
                             min_pts = 2,
                             id_col = "cow",
                             start_col = "start",
                             end_col = "end",
                             bin_col = "bin",
                             intake_col = "intake",
                             dur_col = "duration",
                             tz = "America/Vancouver")
  
  scope_results[[scope]] <- nrow(meals_temp)
}

cat("Meals by eps scope:\n")
for (scope in names(scope_results)) {
  cat(scope, ":", scope_results[[scope]], "meals\n")
}
```

## Labeling Individual Visits

To analyze individual visit patterns within meals, use `meal_label_visits()`:

```{r visit-labeling}
# Label each visit with its meal information
labeled_visits <- meal_label_visits(demo_data,
                                   eps = NULL,
                                   min_pts = 2,
                                   method = "gmm",
                                   id_col = "cow",
                                   start_col = "start",
                                   end_col = "end",
                                   bin_col = "bin",
                                   intake_col = "intake",
                                   dur_col = "duration",
                                   tz = "America/Vancouver")

day1 <- labeled_visits[[1]]

# Check the new columns
head(day1)

# Summary of meal assignments
cat("Visit assignments:\n")
cat("Total visits assigned to meals:", sum(day1$meal_id > 0), "\n")
cat("Outliers (meal_id = 0) on day 1:", sum(day1$meal_id == 0), "\n")
```

## Visualizing Meal Clustering Results

### Timeline Plots

Visualize meals as timeline plots showing when animals fed throughout the day:

- A plot list is returned for each animal on each day. The plot list is organized by animal then date


```{r viz-timeline, fig.width=12, fig.height=8}
# Create timeline plots for meal clusters
meal_plots <- viz_meal_clusters(labeled_visits,
                               id_col = "cow",
                               start_col = "start",
                               color_palette = "Set 3",
                               title_prefix = "Animal",
                               text_size = 10,
                               tz = "America/Vancouver")
# first cow on first day
meal_plots[[1]][[1]]

# first cow on second day
meal_plots[[2]][[2]]

# 10th cow on third day
meal_plots[[10]][[3]]

```

### Individual Animal Plots

For detailed analysis, extract plots for specific animals:

```{r individual-plots, fig.width=10, fig.height=12}
# Get plots for a specific animal
animal_plots <- extract_plots(meal_plots, animals = "5114")

# Combine multiple days for one animal
combined_plots <- combine_animal_plots(meal_plots, 
                                      animal_id = "5114",
                                      plots_per_page = 2,
                                      method = "vertical")

combined_plots[["1"]]
```

### Date-Based Plots

Compare multiple animals on the same date:

```{r date-plots, fig.width=10, fig.height=10}
# Combine multiple animals for one date
date_plots <- combine_date_plots(meal_plots,
                                date = "2020-10-31",
                                plots_per_page = 3,
                                method = "vertical")

date_plots[["1"]]
```

### Custom Visualization Parameters

Experiment with different visualization parameters:

```{r custom-viz, fig.width=12, fig.height=6}
# Custom visualization with different parameters
custom_plots <- viz_meal_clusters(labeled_visits,
                                 id_col = "cow",
                                 start_col = "start",
                                 point_size = 3,
                                 point_alpha = 0.8,
                                 color_palette = "Dark 3",
                                 outlier_color = "red",
                                 title_prefix = "Cow",
                                 text_size = 12,
                                 time_breaks = "2 hours",
                                 tz = "America/Vancouver")

custom_plots
```

## 🎉 Ollie's Success Story!

*"Wow!" Ollie exclaims, looking at all the beautiful meal visualizations. "We've successfully solved the meal mystery! Now we can see clear patterns in how animals actually feed - not just individual visits, but complete feeding events!"*

**What we've accomplished together:**
- ✅ Learned two methods to find optimal meal intervals
- ✅ Understood how DBSCAN groups visits into meals  
- ✅ Experimented with different parameters
- ✅ Created beautiful visualizations of meal patterns
- ✅ Labeled individual visits with their meal information

*"The best part is that you can now use these same techniques with your own animal data!"* 🦦

## 🦦 Ollie's Meal Clustering Cheat Sheet

*"Here's my handy reference guide for all the meal clustering functions! Keep this nearby when working with your own data."*

### 🚀 Ollie's 3-Step Quick Start
```r
# 1. Find the optimal meal interval
eps <- meal_interval(data, method = "gmm", id_col = "cow", 
                    start_col = "start", end_col = "end")

# 2. Cluster visits into meals (one magical step!)
labeled_data <- meal_label_visits(data, id_col = "cow", 
                                 start_col = "start", end_col = "end",
                                 bin_col = "bin", intake_col = "intake", 
                                 dur_col = "duration")

# 3. Create beautiful visualizations
plots <- viz_meal_clusters(labeled_data, id_col = "cow", start_col = "start")
```

### 🔍 Method Comparison Recipes
```r
# Simple percentile approach
eps_pct <- meal_interval(data, method = "percentile", percentile = 0.93)

# Fancy GMM method  
eps_gmm <- meal_interval(data, method = "gmm", use_log_transform = TRUE)

# Conservative "both" approach (uses minimum)
eps_both <- meal_interval(data, method = "both")
```

### 📊 Visualization Menu
```r
# Percentile method visualization
viz_eps_percentile(data, percentile = 0.93, xlim = 200)

# GMM method visualization  
viz_eps_gmm(data, use_log_transform = TRUE, xlim = 10)

# Timeline meal plots
viz_meal_clusters(labeled_data, color_palette = "Set 3", 
                 point_size = 2, title_prefix = "Animal")
```

### ⚙️ Parameter Playground
```r
# Clustering parameter adjustments
cluster_meals(data, eps = 60, min_pts = 3, 
             eps_scope = "one_animal_all_days")

# GMM log transformation tuning
meal_interval(data, method = "gmm", use_log_transform = TRUE,
             log_multiplier = 20, log_offset = 1)

# Set safety bounds for eps
meal_interval(data, lower_bound = 5, upper_bound = 120)
```

### 🎨 Plot Organization Tools
```r
# Extract specific animal or date plots
animal_plots <- extract_plots(plots, animals = c("5114", "4070"))
date_plots <- extract_plots(plots, dates = "2020-10-31")

# Combine multiple plots into pages
combined <- combine_animal_plots(plots, animal_id = "5114", 
                                plots_per_page = 4)
```

---

**🦦 Ollie's Final Words**: *"Remember, the best parameters are the ones that make sense for YOUR data and research questions. Don't be afraid to experiment with different settings until you find what works best!"*

**Next Adventure**: Try these functions with your own animal data! Start with the defaults, then play with the parameters to discover the feeding patterns hidden in your dataset. 🍽️✨ 