---
title: "2. Meal Clustering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Meal Clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(moo4feed)
library(ggplot2)
library(dplyr)
```

## 🚨 Important: Set Your Global Variables First!

*"Before we dive into the meal clustering," Ollie says while adjusting his analyst glasses, "make sure you've configured your global variables to match your data structure. This makes all the functions work smoothly!"*

If you haven't done this yet, run the setup from **Tutorial 1: Data Cleaning**:

```{r global-setup}
# 🎯 IMPORTANT: Set these to match YOUR data structure!
# Copy and modify these lines to match your column names and settings:

set_global_cols(
  id_col = "cow",           # Animal ID column name
  start_col = "start",      # Visit start time column  
  end_col = "end",          # Visit end time column
  bin_col = "bin",          # Bin/feeder ID column
  intake_col = "intake",    # Feed intake amount column
  dur_col = "duration",     # Visit duration column
  tz = "America/Vancouver"  # Your timezone
)

# Check that everything is set correctly:
cat("✅ Global variables set:\n")
cat("ID column:", id_col2(), "\n")
cat("Start time column:", start_col2(), "\n") 
cat("End time column:", end_col2(), "\n")
cat("Bin column:", bin_col2(), "\n")
cat("Intake column:", intake_col2(), "\n")
cat("Duration column:", duration_col2(), "\n")
cat("Timezone:", tz2(), "\n")
```

*"Perfect! Now all our functions will automatically use the right column names for your data!"* 🦦

## Ollie's Meal Clustering Journey 🦦

*Our favorite data otter Ollie is back! This time, he's puzzled by something he noticed while watching the cows at the research farm...*

"I've been tracking when each cow visits the feeders," Ollie says, adjusting his tiny data analyst glasses, "but I realized something important - just because a cow makes two separate visits doesn't mean they're having two separate meals! Sometimes they just take a quick break and come back."

"So I thought, what if we could group visits into meals and visualize each cow's breakfast, lunch and dinner?"

### The Meal Clustering Challenge

Individual feeding visits tell only part of the story. To understand animal feeding behavior, we need to group these visits into meaningful **meals** - periods of sustained feeding activity separated by longer breaks.

Ollie's goal is to help you:
1. **Analyze temporal patterns** - Look at time gaps between consecutive visits
2. **Find natural breakpoints** - Identify optimal time thresholds that separate within-meal from between-meal visit gaps  
3. **Group visits intelligently** - Use clustering algorithms to group nearby visits into meals
4. **Summarize behavior** - Create meal-level summaries for analysis

*"Let's solve this puzzle together!"* 🦦

## Understanding Inter-Visit Gaps

*"First," says Ollie, "we need to understand the pattern of time gaps between feeding visits. Are there short gaps (within meals) and long gaps (between meals)?"*

Let's help Ollie examine some example data:

```{r data-prep}
# Use a subset of animals for demonstration
demo_data <- clean_feed

# Quick look at our data
cat("Cows on the second day:", unique(demo_data[[2]]$cow), "\n")
cat("Dates with data:", names(demo_data), "\n")
```

## Method 1: Percentile-Based Interval Selection

The simplest approach uses percentiles of inter-visit gaps to determine the optimal meal interval (eps).

### How It Works

The percentile method:
1. Calculates gaps between consecutive visits for each animal within each day
2. Takes a high percentile (default 93rd) as the threshold
3. Visits separated by less than this threshold belong to the same meal

```{r percentile-method}
# Calculate optimal interval using percentile method without restricting the 
# interval has to be between the lower_bound and upper_bound (default is 5 to 60 minutes)
eps_percentile <- meal_interval(data = demo_data, 
                               method = "percentile", 
                               percentile = 0.93,
                               lower_bound = NULL, # default is 5 minutes
                               upper_bound = NULL, # default is 60 minutes
                               use_log_transform = TRUE,
                               log_multiplier = 20,
                               log_offset = 1,
                               id_col = id_col2(), # you don't need to specify this and the 3 other column names, the function will use the global variables
                               start_col = start_col2(), 
                               end_col = end_col2(),
                               tz = tz2())

cat("Optimal interval (93rd percentile):", round(eps_percentile, 2), "minutes\n")
```

### Visualizing the Percentile Method

- Now let's restrict the interval to be between 5 and 60 minutes and visualize the distribution of the gaps.

```{r viz-percentile, fig.width=10, fig.height=6}
# Visualize gap distribution with percentile-based eps
# 🎯 TRY THIS: Change these percentile values!
# Try changing the percentile to 0.75, 0.85, 0.95, 0.99 and see how it affects the threshold!
p1 <- viz_eps_percentile(data = demo_data, 
                        percentile = 0.93,  # 🎯 Change this value!
                        lower_bound = 5, 
                        upper_bound = 60,
                        bins = 100,
                        colors = grDevices::hcl.colors(2, "Set 3"),
                        title_prefix = "Distribution of time gap between visits & \noptimal meal interval (eps)",
                        xlim = 200)

p1
```

## Method 2: Gaussian Mixture Model (GMM) Approach

*"Wait!" Ollie's eyes light up. "I just learned about a fancier statistical method! What if we fit mathematical curves to our gap data to find where short gaps and long gaps naturally separate?"*

### How Ollie's GMM Method Works

The GMM method is like fitting two bell curves to our data:
1. Fits a 2-component Gaussian mixture model to inter-visit gaps
2. Identifies two populations: **within-meal gaps** (short) and **between-meal gaps** (long)
3. Finds the intersection point between these distributions as the optimal threshold
4. Can use log-transformation to better handle skewed gap distributions

*"It's like finding where two waves cross!"* 🌊

```{r gmm-method}
# Calculate optimal interval using GMM method
eps_gmm <- meal_interval(data = demo_data, 
                        method = "gmm", 
                        percentile = 0.93,
                        lower_bound = NULL,
                        upper_bound = NULL,
                        use_log_transform = TRUE,  # 🎯 Try changing this to FALSE!
                        log_multiplier = 20,       # 🎯 Try different values: 1, 10, 20
                        log_offset = 1)          # 🎯 Try different values: 0.5, 2

cat("Optimal interval (GMM):", round(eps_gmm, 2), "minutes\n")
```

### Visualizing the GMM Method

```{r viz-gmm, fig.width=10, fig.height=6}
# Visualize gap distribution with GMM fit
p2 <- viz_eps_gmm(data = demo_data, 
                  lower_bound = NULL,
                  upper_bound = NULL,
                  bins = 100,
                  colors = grDevices::hcl.colors(4, "Set 3"),
                  title_prefix = "Distribution of time gap between visits \n& GMM-based meal interval (eps)",
                  show_components = TRUE,
                  xlim = 10,
                  use_log_transform = TRUE,  # 🎯 Try changing this to FALSE!
                  log_multiplier = 20,       # 🎯 Try different values!
                  log_offset = 1)           # 🎯 Try different values!

p2
```


### 🎯 Experiment with Log Transformations in GMM!

*"The log transformation has some knobs we can twist!" Ollie says excitedly. "Let's see how different parameters affect our results!"*

**🔄 Interactive Challenge**: Try different log transformation parameters:

```{r gmm-parameters}
# 🎯 TRY THIS: Change these parameter combinations!
# Try different multipliers: 5, 10, 20, 30, 50
# Try different offsets: 0.5, 1, 2
log_params <- list(
  list(multiplier = 1, offset = 1),    # 🎯 Change these values!
  list(multiplier = 20, offset = 1),    # 🎯 Change these values!
  list(multiplier = 50, offset = 1)     # 🎯 Change these values!
)

for (i in seq_along(log_params)) {
  eps_val <- meal_interval(data = demo_data, 
                          method = "gmm", 
                          percentile = 0.93,
                          lower_bound = 5,
                          upper_bound = 60,
                          use_log_transform = TRUE,
                          log_multiplier = log_params[[i]]$multiplier,  # 🎯 Your input here!
                          log_offset = log_params[[i]]$offset)         # 🎯 Your input here!
  
  cat("Log params (mult=", log_params[[i]]$multiplier, ", offset=", log_params[[i]]$offset, 
      "): eps =", round(eps_val, 2), "minutes\n")
}
```

### Visualize different Log Transformations

Log transformation often improves GMM fitting for highly skewed gap distributions:

```{r gmm-comparison, fig.width=12, fig.height=8}
# Compare GMM with and without log transformation
p_gmm_no_log <- viz_eps_gmm(demo_data, 
                            lower_bound = NULL,
                            upper_bound = NULL,
                            bins = 100,
                            colors = grDevices::hcl.colors(4, "Set 3"),
                            title_prefix = "Distribution of time gap between visits \n& GMM-based meal interval (eps)",
                            show_components = TRUE,
                            use_log_transform = FALSE,
                            xlim = 20)
p_gmm_no_log

# with basic log transformation: log(x + 1) where x is the gap in minutes
p_gmm_log <- viz_eps_gmm(demo_data, 
                        lower_bound = NULL,
                        upper_bound = NULL,
                        bins = 100,
                        colors = grDevices::hcl.colors(4, "Set 3"),
                        title_prefix = "Distribution of time gap between visits \n& GMM-based meal interval (eps)",
                        show_components = TRUE,
                        use_log_transform = TRUE,
                        log_multiplier = 1,
                        log_offset = 1,
                        xlim = 10)
p_gmm_log

# with customzied log transformation: log(20x + 1) where x is the gap in minutes
p_gmm_log_20 <- viz_eps_gmm(demo_data, 
                            lower_bound = NULL,
                            upper_bound = NULL,
                            bins = 100,
                            colors = grDevices::hcl.colors(4, "Set 3"),
                            title_prefix = "Distribution of time gap between visits \n& GMM-based meal interval (eps)",
                            show_components = TRUE,
                            use_log_transform = TRUE,
                            log_multiplier = 20,
                            log_offset = 1,
                            xlim = 10)
p_gmm_log_20

```

**💡 Ollie's Tip**: *"Different parameters can give different results. The key is finding what works best for your specific data!"*

## DBSCAN: Ollie's Clustering Algorithm

*"Now comes the exciting part!" Ollie grins. "Once we know our optimal interval, we need to actually group the visits into meals. I'll use something called DBSCAN - it's like finding dense neighborhoods of visits!"*

### How Ollie's DBSCAN Works

DBSCAN is like Ollie looking for "busy neighborhoods of visits":

1. **Core points**: Visits with enough nearby friends (at least `min_pts` neighbors within `eps` time)
2. **Border points**: Visits near a busy neighborhood but not busy themselves
3. **Noise points**: Lonely visits that don't belong to any meal (isolated visits)
4. **Clusters**: Groups of connected busy neighborhoods = **MEALS!**

For meal clustering:
- **Points** = feeding visit start times 🕐
- **Distance** = time difference between the end time of the previous visit and the start time of the current visit ⏱️
- **eps** = maximum time gap within a meal ⏰
- **min_pts** = minimum visits to form a dense meal cluster 🍽️

*"It's like connecting the dots, but only if the dots are close enough together!"* 🦦

## Clustering Visits into Meals

Now let's cluster our feeding visits into meals using different parameters:

### Basic Clustering

```{r basic-clustering}
# Cluster meals with default parameters
# 🎯 TRY THIS: Change these min_pts values!
# Try: c(1, 2, 3, 4) or c(2, 3, 5, 7) or even c(1, 2, 4, 8)
# Rule of thumb: min_pts >= D + 1 where D is the number of dimensions of the data.
# So for us min_pts = 1 + 1 = 2
meals_basic <- cluster_meals(data = demo_data,
                            eps = NULL,  # Auto-determine using GMM
                            min_pts = 2,  # 🎯 Try changing this to 3, 4, or 5!
                            method = "gmm",
                            percentile = 0.93,
                            eps_scope = "all_animals",
                            lower_bound = NULL,
                            upper_bound = NULL,
                            use_log_transform = TRUE,
                            log_multiplier = 20,
                            log_offset = 1)

# Look at the results
head(meals_basic)
cat("Total meals identified:", nrow(meals_basic), " in 2 days\n")
cat("Meals per animal:\n")
head(table(meals_basic$cow, meals_basic$date))
```


### Different eps Scopes

You can calculate optimal eps using different data scopes:
- **all_animals** = calculate a universal optimal eps using visit gaps for all animals from all days
- **one_animal_all_days** = Customize eps for each individual animal, calculate eps based on every animal across all days
- **one_animal_single_day** = Customize eps for each individual and different days, calculate eps based on each animal on each day

```{r eps-scopes}
# Compare different eps calculation scopes
scopes <- c("all_animals", "one_animal_all_days", "one_animal_single_day")
scope_results <- list()

for (scope in scopes) {
  meals_temp <- cluster_meals(data = demo_data,
                             eps = NULL,
                             min_pts = 2,
                             method = "gmm",
                             percentile = 0.93,
                             eps_scope = scope,  # 🎯 This changes in the loop!
                             lower_bound = NULL,
                             upper_bound = NULL,
                             use_log_transform = TRUE,
                             log_multiplier = 20,
                             log_offset = 1)
  
  scope_results[[scope]] <- nrow(meals_temp)
}

cat("Meals by eps scope:\n")
for (scope in names(scope_results)) {
  cat(scope, ":", scope_results[[scope]], "meals\n")
}
```

## Labeling Individual Visits

- The `meal_label_visits()` function respect your input data structure. If your input data is a list of dataframes, it will return a labeled list of dataframes; if your input data is a dataframe, it will return a labeled dataframe.

- To analyze individual visit patterns within meals, use `meal_label_visits()`.

```{r visit-labeling}
# Label each visit with its meal information
labeled_visits <- meal_label_visits(data = demo_data,
                                   eps = NULL,
                                   min_pts = 2,
                                   method = "gmm",
                                   percentile = 0.93,
                                   eps_scope = "all_animals",
                                   lower_bound = NULL,
                                   upper_bound = NULL,
                                   use_log_transform = TRUE,
                                   log_multiplier = 20,
                                   log_offset = 1)

day1 <- labeled_visits[[1]]

# Check the new columns
head(day1)

# Summary of meal assignments
cat("Visit assignments:\n")
cat("Total visits assigned to meals:", sum(day1$meal_id > 0), "\n")
cat("Feeding visits that are not assigned to meals (outliers,meal_id = 0) on day 1:", sum(day1$meal_id == 0), "\n")
```

## Visualizing Meal Clustering Results

### Timeline Plots

Visualize meals as timeline plots showing when animals fed throughout the day:

- A plot list is returned for each animal on each day. The plot list is organized by animal then date

```{r viz-timeline, fig.width=12, fig.height=8}
# Create timeline plots for meal clusters
meal_plots <- viz_meal_clusters(data = labeled_visits,
                               point_size = 2,
                               point_alpha = 0.7,
                               ncol_facet = 1,
                               date_format = "%Y-%m-%d",
                               time_breaks = "4 hours",
                               time_labels = "%H",
                               color_palette = "Set 3",  # 🎯 Try "Dark 2", "Pastel 1"!
                               outlier_color = "grey50",
                               title_prefix = "Cow",  # 🎯 Try "Animal" or ""!
                               text_size = 10,
                               title_size = NULL)
# first digit means sequence of the animal, second digit means sequence of the day
# first cow on first day
meal_plots[[1]][[1]]

# first cow on second day
meal_plots[[2]][[2]]

# 10th cow on second day
meal_plots[[10]][[2]]

```

### Individual Animal Plots

For detailed analysis, extract plots for specific animals:

```{r individual-plots, fig.width=10, fig.height=12}
# Get plots for a specific animal on specific day
animal_plot <- extract_plots(meal_plots, animals = "6084", dates = "2020-10-31")
animal_plot

# Combine multiple days for one animal
combined_plots <- combine_animal_plots(meal_plots, 
                                      animal_id = "7022",
                                      plots_per_page = 3,
                                      method = "vertical")

combined_plots[[1]]
```

### Date-Based Plots

Compare multiple animals on the same date:

```{r date-plots, fig.width=10, fig.height=10}
# Combine multiple animals for one date
date_plots <- combine_date_plots(meal_plots,
                                date = "2020-10-31",
                                plots_per_page = 3,
                                method = "vertical")

date_plots[["1"]]
```

### Custom Visualization Parameters

Experiment with different visualization parameters:

```{r custom-viz, fig.width=12, fig.height=6}
# Custom visualization with different parameters
custom_plots <- viz_meal_clusters(data = labeled_visits,
                                 point_size = 2,        # 🎯 Try different sizes: 1, 2, 4!
                                 point_alpha = 0.8,     # 🎯 Try: 0.5, 0.9, 1.0!
                                 ncol_facet = 1,
                                 date_format = "%Y-%m-%d",
                                 time_breaks = "2 hours",  # 🎯 Try "1 hour", "6 hours"!
                                 time_labels = "%H",
                                 color_palette = "Dark 2",  # 🎯 Try different palettes!
                                 outlier_color = "red",     # 🎯 Try different colors!
                                 title_prefix = "Cow",      # 🎯 Try "Animal" or ""!
                                 text_size = 12,           # 🎯 Try different sizes!
                                 title_size = NULL)

custom_plots[[2]][[1]]
```

## 🎉 Ollie's Success Story!

*"Wow!" Ollie exclaims, looking at all the beautiful meal visualizations. "We've successfully solved the meal mystery! Now we can see clear patterns in how animals actually feed - not just individual visits, but when they have breakfast, afternoon snacks, and more!"*

*"The best part is that you can now use these same techniques with your own animal data!"* 🦦

## 🦦 Meal Clustering Cheat Sheet

*"Here's my handy reference guide for all the meal clustering functions! Copy and paste these code blocks to work with your own data."*

```r
#' ============================================================================
#' 🦦 OLLIE'S MEAL CLUSTERING CHEAT SHEET
#' ============================================================================
#' Copy and modify these code blocks for your own analysis!

# ---- SETUP: Global Variables (REQUIRED FIRST!) ----
library(moo4feed)
library(ggplot2)
library(dplyr)

# Set up your column names and timezone (modify these!)
set_global_cols(
  id_col = "cow",           # Your animal ID column
  start_col = "start",      # Visit start time column  
  end_col = "end",          # Visit end time column
  bin_col = "bin",          # Bin/feeder ID column
  intake_col = "intake",    # Feed intake amount column
  dur_col = "duration",     # Visit duration column
  tz = "America/Vancouver"  # Your timezone
)

# ---- STEP 1: Find Optimal Meal Interval ----
# Method 1: Simple percentile approach
eps_percentile <- meal_interval(
  data = your_data, 
  method = "percentile", 
  percentile = 0.93,                    # Try: 0.80, 0.90, 0.95, 0.99
  lower_bound = 5, 
  upper_bound = 60,
  id_col = id_col2(), 
  start_col = start_col2(), 
  end_col = end_col2(),
  tz = tz2()
)

# Method 2: Gaussian Mixture Model (more sophisticated)
eps_gmm <- meal_interval(
  data = your_data, 
  method = "gmm", 
  percentile = 0.93,
  lower_bound = 5, 
  upper_bound = 60,
  use_log_transform = TRUE,             # Try: FALSE
  log_multiplier = 20,                  # Try: 10, 30, 50
  log_offset = 1,                       # Try: 0.5, 2
  id_col = id_col2(), 
  start_col = start_col2(), 
  end_col = end_col2(),
  tz = tz2()
)

# Method 3: Conservative approach (uses both methods, takes minimum)
eps_both <- meal_interval(
  data = your_data, 
  method = "both", 
  percentile = 0.93,
  id_col = id_col2(), 
  start_col = start_col2(), 
  end_col = end_col2(),
  tz = tz2()
)

# ---- STEP 2: Visualize Gap Distributions ----
# Visualize percentile method
p1 <- viz_eps_percentile(
  data = your_data, 
  percentile = 0.93,                    # Try different values!
  lower_bound = 5, 
  upper_bound = 60,
  xlim = 200,                           # Adjust for your data range
  id_col = id_col2(), 
  start_col = start_col2(), 
  end_col = end_col2(),
  tz = tz2()
)

# Visualize GMM method  
p2 <- viz_eps_gmm(
  data = your_data, 
  use_log_transform = TRUE,             # Try: FALSE
  xlim = 10,                            # Use smaller values for log transform
  show_components = TRUE,               # Try: FALSE
  id_col = id_col2(), 
  start_col = start_col2(), 
  end_col = end_col2(),
  tz = tz2()
)

# ---- STEP 3: Cluster Visits into Meals ----
# - Step 1 + 2 are optional, it's designed to help you find the optimal interval (eps) 
#   or find the best automatic method of identifyingoptimal eps for the clustering.
# - You can skip step 1 + 2 and just use the 2 functions below if you are confident.
# Option A: Just get meal summaries
meals <- cluster_meals(
  data = your_data,
  eps = NULL,                           # Auto-determine, or set specific value
  min_pts = 2,                          # Try: 3, 4, 5 for stricter clustering
  method = "gmm",                       # Try: "percentile", "both"
  eps_scope = "all_animals",            # Try: "one_animal_all_days", "one_animal_single_day"
  lower_bound = 5, 
  upper_bound = 60,
  id_col = id_col2(),
  start_col = start_col2(),
  end_col = end_col2(),
  bin_col = bin_col2(),
  intake_col = intake_col2(),
  dur_col = duration_col2(),
  tz = tz2()
)

# Option B: Label individual visits with meal info (recommended!)
labeled_visits <- meal_label_visits(
  data = your_data,
  eps = NULL,                           # Auto-determine optimal interval
  min_pts = 2,                          # Minimum visits to form a meal
  method = "gmm",                       # Clustering method
  eps_scope = "all_animals",            # Scope for eps calculation
  id_col = id_col2(),
  start_col = start_col2(),
  end_col = end_col2(),
  bin_col = bin_col2(),
  intake_col = intake_col2(),
  dur_col = duration_col2(),
  tz = tz2()
)

# ---- STEP 4: Visualize Meal Patterns ----
# Create timeline plots showing meals
meal_plots <- viz_meal_clusters(
  data = labeled_visits,
  point_size = 2,                       # Try: 1, 3, 4
  point_alpha = 0.7,                    # Try: 0.5, 0.8, 1.0
  color_palette = "Set 3",              # Try: "Dark 2", "Pastel 1", "Set 1"
  outlier_color = "grey50",             # Try: "red", "black"
  title_prefix = "Animal",              # Try: "Cow", ""
  text_size = 12,                       # Try: 10, 14, 16
  time_breaks = "4 hours",              # Try: "2 hours", "6 hours"
  id_col = id_col2(),
  start_col = start_col2(),
  tz = tz2()
)

# ---- STEP 5: Extract and Combine Specific Plots ----
# Get plots for specific animals
animal_plots <- extract_plots(
  plot_list = meal_plots, 
  animals = c("5114", "4070"),          # Your animal IDs
  dates = NULL                          # All dates, or specify: c("2020-10-31")
)

# Get plots for specific dates
date_plots <- extract_plots(
  plot_list = meal_plots,
  animals = NULL,                       # All animals
  dates = "2020-10-31"                  # Your specific date
)

# Combine multiple days for one animal
combined_plots <- combine_animal_plots(
  plot_list = meal_plots, 
  animal_id = "5114",                   # Your animal ID
  plots_per_page = 4,                   # Number of plots per page
  method = "vertical"                   # Try: "grid"
)

# Combine multiple animals for one date
date_combined <- combine_date_plots(
  plot_list = meal_plots,
  date = "2020-10-31",                  # Your specific date
  plots_per_page = 3,
  method = "vertical"
)

# ---- BONUS: Quick Analysis ----
# Check meal summary statistics
summary(meals$meal_duration)           # Meal durations in seconds
summary(meals$visit_count)             # Visits per meal
summary(meals$total_intake)            # Intake per meal
table(meals$cow, meals$date)           # Meals per animal, per day

# Check labeling success on the first day
table(labeled_visits[[1]]$meal_id == 0)    # TRUE = outliers, FALSE = assigned to meals

#' ============================================================================
#' 🦦 That's it! You're now ready to analyze meal patterns in your data!
#' ============================================================================
```

---

**🦦 Ollie's Final Words**: *"Remember, the best parameters are the ones that make sense for YOUR data and research questions. Don't be afraid to experiment with different settings until you find what works best!"*

**Next Adventure**: Try these functions with your own animal data! Start with the defaults, then play with the parameters to discover the feeding patterns hidden in your dataset. 🍽️✨ 