---
title: "6. Non-Nutritive Visit Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{6. Non-Nutritive Visit Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(moo4feed)
library(ggplot2)
library(dplyr)
```

## 1. 🚨 Important: Set Your Global Variables First

Before analyzing non-nutritive visit patterns, configure global variables to match your data structure:

```{r global-setup}
# Configure global variables for your data structure
set_global_cols(
  id_col = "cow",           # Animal ID column name
  start_col = "start",      # Visit start time column
  end_col = "end",          # Visit end time column
  bin_col = "bin",          # Bin/feeder ID column
  intake_col = "intake",    # Feed intake amount column
  dur_col = "duration",     # Visit duration column
  tz = "America/Vancouver"  # Your timezone
)

```

## 2. Introduction to Non-Nutritive Visits

In nature, animals explore their environment while foraging. In the barn, we have observed that not every bin visit result in the cow actually feeding. Sometimes they visit a feeder, see that feed is available, but choose not to eat. We refer to those as non-nutritive visits, which potentially reflects exploratory behaviour.

## 3. Prerequisites

This tutorial assumes completion of previous data processing steps in [**Tutorial 1: Data Cleaning**](data_cleaning.html).

## 4. Data Preparation

```{r data-prep}
# Load cleaned example data
data(clean_feed)

# If you're using your own data from previous tutorials, use this instead:
# clean_feed <- your_cleaned_feed_data     # From your cleaning results
```

## 5. Understanding Non-Nutritive Visits

A **non-nutritive visit** is defined as when:

* The animal visits a bin
* Feed is available in the bin (start weight > calibration error)
* The animal consumes little or no feed (intake ≤ calibration error)

The **calibration error** is the measurement threshold below which values are considered zero due to equipment sensitivity (typically 0.5 kg for feed bins).

### Sorting Results

Both `calculate_non_nutritive_visits()` and `calculate_no_feed_visits()` support an optional `sort` parameter to order results by visit frequency:

* `sort = 0` (default) - No sorting
* `sort = 1` - Ascending order (lowest to highest visits)  
* `sort = -1` - Descending order (highest to lowest visits)

This feature helps quickly identify animals with the most (or least) visits of each type.

### Calculate Non-Nutritive Visits

```{r non-nutritive-analysis}
# Create quality control configuration with calibration error
my_qc_config <- qc_config(
  calibration_error = 0.5    # Equipment measurement threshold (kg)
)

# Calculate non-nutritive visits for each animal on each day. 
non_nutritive <- calculate_non_nutritive_visits(
  data = clean_feed,           # Our cleaned feed data
  cfg = my_qc_config           # Configuration with calibration error
)

# Examine the first day's results (unsorted)
cat("Non-nutritive visits on first day (unsorted):\n")
head(non_nutritive[[1]])

```

### Using the Sort Parameter

```{r sort-examples}
# Sort by highest non-nutritive visits first (descending)
non_nutritive_desc <- calculate_non_nutritive_visits(
  data = clean_feed,
  cfg = my_qc_config,
  sort = -1    # Sort descending (highest first)
)

cat("\n Animals with MOST non-nutritive visits (top 5):\n")
head(non_nutritive_desc[[1]], 5)

# Sort by lowest non-nutritive visits first (ascending)
non_nutritive_asc <- calculate_non_nutritive_visits(
  data = clean_feed,
  cfg = my_qc_config,
  sort = 1     # Sort ascending (lowest first)
)

cat("\n Animals with LEAST non-nutritive visits (top 5):\n")
head(non_nutritive_asc[[1]], 5)

```

## 6. Understanding No-Feed Visits

A **no-feed visit** (or empty bin visit) occurs when:

* The animal visits a feeding bin
* No feed is available in the bin (start weight ≤ calibration error)
* The animal cannot consume anything (intake ≤ calibration error)

These visits indicate animals are checking bins that are already empty, which may reflect:

* The animal having lots of empty bin visits may indicate that they are disadvantaged, because they can only eat the "leftovers" after others have finished eating.
* High feeding competition (bins emptied quickly)
* Feed management issues (e.g., we may need to adjust the feed delivery timing or amount)

### Calculate No-Feed Visits

```{r no-feed-analysis}
# Calculate visits to empty bins for each animal on each day
no_feed <- calculate_no_feed_visits(
  data = clean_feed,           # Our cleaned feed data
  cfg = my_qc_config           # Configuration with calibration error
)

# Examine the first day's results (unsorted)
cat("No-feed visits on first day (unsorted):\n")
head(no_feed[[1]])

# Sort by highest empty bin visits (descending) to identify animals checking empty bins most
no_feed_desc <- calculate_no_feed_visits(
  data = clean_feed,
  cfg = my_qc_config,
  sort = -1    # Sort descending (highest first)
)

cat("\n Animals checking EMPTY bins most often (top 5):\n")
head(no_feed_desc[[1]], 5)

```

### Visualize Non-Nutritive and No-Feed Patterns

```{r viz-non-nutritive-bar, fig.width=10, fig.height=6}
# Use first day's data and get animals with most non-nutritive visits
top_nn <- head(non_nutritive_desc[[1]], 50)

ggplot(top_nn, aes(x = reorder(cow, number_of_non_nutritive_visits), y = number_of_non_nutritive_visits)) +
  geom_bar(stat = "identity", fill = "coral", alpha = 0.7) +
  coord_flip() +
  labs(
    title = "Frequency of Non-Nutritive Visits",
    subtitle = "Non-nutritive visits: Feed available but animal did not eat",
    x = "Animal ID",
    y = "Total Non-Nutritive Visits"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11)
  )
```

```{r viz-no-feed-bar, fig.width=10, fig.height=6}
# Use first day's data and get animals checking empty bins
top_empty <- head(no_feed_desc[[1]], 50)

ggplot(top_empty, aes(x = reorder(cow, number_of_visits_when_no_feed), y = number_of_visits_when_no_feed)) +
  geom_bar(stat = "identity", fill = "olivedrab3", alpha = 0.7) +
  coord_flip() +
  labs(
    title = "Frequency of Empty Bin Visits",
    subtitle = "Empty bin visits: No feed available in the visited bin",
    x = "Animal ID",
    y = "Total Empty Bin Visits"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11)
  )
```

**Note:** Many animals have no empty bin visits, so they are excluded from the plot.

## 7. Code Cheatsheet

```{r cheatsheet, eval=FALSE}
#' Copy and modify these code blocks for your own analysis!

# ---- SETUP: Global Variables (REQUIRED FIRST!) ----
library(moo4feed)
library(ggplot2)
library(dplyr)

# Set up your column names and timezone (modify these!)
set_global_cols(
  id_col = "cow",           # Your animal ID column
  start_col = "start",      # Visit start time column
  end_col = "end",          # Visit end time column
  bin_col = "bin",          # Bin/feeder ID column
  intake_col = "intake",    # Feed intake amount column
  dur_col = "duration",     # Visit duration column
  tz = "America/Vancouver"  # Your timezone
)

# ---- STEP 1: Load Your Data ----
# Load your cleaned data
data(clean_feed)

# Or use your own cleaned data from previous tutorials:
# clean_feed <- your_cleaned_feed_data

# ---- STEP 2: Create QC Configuration ----
my_qc_config <- qc_config(
  calibration_error = 0.5    # Equipment measurement threshold (kg)
)

# ---- STEP 3: Calculate Non-Nutritive Visits ----
# Basic calculation (unsorted)
non_nutritive <- calculate_non_nutritive_visits(
  data = clean_feed,
  cfg = my_qc_config
)

# View first day results
head(non_nutritive[[1]])

# Sort by highest visits first (descending)
non_nutritive_desc <- calculate_non_nutritive_visits(
  data = clean_feed,
  cfg = my_qc_config,
  sort = -1
)

head(non_nutritive_desc[[1]], 5)

# Sort by lowest visits first (ascending)
non_nutritive_asc <- calculate_non_nutritive_visits(
  data = clean_feed,
  cfg = my_qc_config,
  sort = 1
)

head(non_nutritive_asc[[1]], 5)

# ---- STEP 4: Calculate No-Feed Visits ----
# Basic calculation (unsorted)
no_feed <- calculate_no_feed_visits(
  data = clean_feed,
  cfg = my_qc_config
)

# View first day results
head(no_feed[[1]])

# Sort by highest empty bin visits (descending)
no_feed_desc <- calculate_no_feed_visits(
  data = clean_feed,
  cfg = my_qc_config,
  sort = -1
)

head(no_feed_desc[[1]], 5)

# ---- STEP 5: Visualize Results ----
# Non-nutritive visits bar plot (first day example)
top_nn <- head(non_nutritive_desc[[1]], 50)

ggplot(top_nn, aes(x = reorder(cow, number_of_non_nutritive_visits), y = number_of_non_nutritive_visits)) +
  geom_bar(stat = "identity", fill = "coral", alpha = 0.7) +
  coord_flip() +
  labs(
    title = "Frequency of Non-Nutritive Visits",
    subtitle = "Non-nutritive visits: Feed available but animal did not eat",
    x = "Animal ID",
    y = "Total Non-Nutritive Visits"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11)
  )

# Empty bin visits bar plot (first day example)
top_empty <- head(no_feed_desc[[1]], 50)

ggplot(top_empty, aes(x = reorder(cow, number_of_visits_when_no_feed), y = number_of_visits_when_no_feed)) +
  geom_bar(stat = "identity", fill = "olivedrab3", alpha = 0.7) +
  coord_flip() +
  labs(
    title = "Frequency of Empty Bin Visits",
    subtitle = "Empty bin visits: No feed available in the visited bin",
    x = "Animal ID",
    y = "Total Empty Bin Visits"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 8),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11)
  )
```

---
