---
title: "6. Non-Nutritive Visit Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{5. Non-Nutritive Visit Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(moo4feed)
library(ggplot2)
library(dplyr)
```

## 1. 🚨 Important: Set Your Global Variables First!

Before analyzing non-nutritive visit patterns, configure global variables to match your data structure:

```{r global-setup}
# Configure global variables for your data structure
set_global_cols(
  id_col = "cow",           # Animal ID column name
  start_col = "start",      # Visit start time column
  end_col = "end",          # Visit end time column
  bin_col = "bin",          # Bin/feeder ID column
  intake_col = "intake",    # Feed intake amount column
  dur_col = "duration",     # Visit duration column
  tz = "America/Vancouver"  # Your timezone
)

# Verify configuration
cat("✅ Global variables configured:\n")
cat("ID column:", id_col2(), "\n")
cat("Start time column:", start_col2(), "\n")
cat("End time column:", end_col2(), "\n")
cat("Bin column:", bin_col2(), "\n")
cat("Intake column:", intake_col2(), "\n")
cat("Duration column:", duration_col2(), "\n")
cat("Timezone:", tz2(), "\n")
```

## 2. Introduction to Non-Nutritive Visits

🦦 Ollie the Otter explains: "Not every bin visit results in eating! Sometimes animals visit a feeder, see that feed is available, but choose not to eat. These are called non-nutritive visits. Understanding this behavior helps researchers identify selective feeding patterns, food preferences, and potential health issues."

Non-nutritive visits can reveal important behavioral insights:

- **Feed selectivity**: Animals may reject certain feed types or quality
- **Social dynamics**: Animals may visit bins to assess feeding opportunities without committing
- **Health monitoring**: Changes in non-nutritive visit patterns may indicate illness or stress
- **Resource competition**: High non-nutritive visits may suggest animals are frequently displaced before they can eat

### What We'll Learn

This tutorial demonstrates how to:

> 1. **Calculate non-nutritive visits** - Identify when animals visit bins with feed available but don't eat
> 2. **Calculate no-feed visits** - Identify when animals visit empty bins
> 3. **Analyze patterns** - Understand individual differences in feeding selectivity

## 3. Prerequisites

This tutorial assumes completion of previous data processing steps in **Tutorial 1: Data Cleaning**

## 4. Data Preparation

```{r data-prep}
# Load cleaned example data
data(clean_feed)

# If you're using your own data from previous tutorials, use this instead:
# clean_feed <- your_cleaned_feed_data     # From your cleaning results

# Quick peek at our data structure
cat("Feed data structure:\n")
head(clean_feed[[1]], 3)  # First day, first 3 rows

cat("\nTotal days of data:", length(clean_feed), "\n")
cat("Number of animals on first day:", length(unique(clean_feed[[1]]$cow)), "\n")
```

## 5. Understanding Non-Nutritive Visits

A **non-nutritive visit** occurs when:
- The animal visits a feeding bin
- Feed is available in the bin (start weight > calibration error)
- The animal consumes little or no feed (intake ≤ calibration error)

The **calibration error** is the measurement threshold below which values are considered zero due to equipment sensitivity (typically 0.5 kg for feed bins).

### Calculate Non-Nutritive Visits

```{r non-nutritive-analysis}
# Create quality control configuration with calibration error
my_qc_config <- qc_config(
  calibration_error = 0.5    # Equipment measurement threshold (kg)
)

# Calculate non-nutritive visits for each animal on each day
non_nutritive <- calculate_non_nutritive_visits(
  data = clean_feed,           # Our cleaned feed data
  cfg = my_qc_config           # Configuration with calibration error
)

# Examine the first day's results
cat("Non-nutritive visits on first day:\n")
head(non_nutritive[[1]])

# Summary statistics
cat("\nSummary of non-nutritive visits (first day):\n")
summary(non_nutritive[[1]]$number_of_non_nutritive_visits)
```

## 6. Understanding No-Feed Visits

A **no-feed visit** (or empty bin visit) occurs when:
- The animal visits a feeding bin
- No feed is available in the bin (start weight ≤ calibration error)
- The animal cannot consume anything (intake ≤ calibration error)

These visits indicate animals are checking bins that are already empty, which may reflect:
- High feeding competition (bins emptied quickly)
- Poor feed distribution timing
- Exploration behavior

### Calculate No-Feed Visits

```{r no-feed-analysis}
# Calculate visits to empty bins for each animal on each day
no_feed <- calculate_no_feed_visits(
  data = clean_feed,           # Our cleaned feed data
  cfg = my_qc_config           # Configuration with calibration error
)

# Examine the first day's results
cat("No-feed visits on first day:\n")
head(no_feed[[1]])

# Summary statistics
cat("\nSummary of no-feed visits (first day):\n")
summary(no_feed[[1]]$number_of_visits_when_no_feed)
```

## 7. Comparing Non-Nutritive and No-Feed Patterns

Let's combine both metrics to understand the full picture of feeding behavior:

```{r combined-analysis}
# Combine non-nutritive and no-feed data for first day
combined_day1 <- non_nutritive[[1]] |>
  dplyr::full_join(
    no_feed[[1]],
    by = "cow"
  ) |>
  dplyr::mutate(
    number_of_non_nutritive_visits = tidyr::replace_na(number_of_non_nutritive_visits, 0),
    number_of_visits_when_no_feed = tidyr::replace_na(number_of_visits_when_no_feed, 0)
  )

# Display animals with highest non-nutritive visits
cat("Animals with most non-nutritive visits (first day):\n")
combined_day1 |>
  dplyr::arrange(dplyr::desc(number_of_non_nutritive_visits)) |>
  head(5)

# Display animals with highest no-feed visits
cat("\nAnimals with most no-feed visits (first day):\n")
combined_day1 |>
  dplyr::arrange(dplyr::desc(number_of_visits_when_no_feed)) |>
  head(5)
```

### Visualize the Relationship

```{r visualization, fig.width=6, fig.height=5}
# Create scatter plot showing relationship between visit types
ggplot(combined_day1, aes(x = number_of_non_nutritive_visits,
                          y = number_of_visits_when_no_feed)) +
  geom_point(alpha = 0.6, size = 3, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "coral", linetype = "dashed") +
  labs(
    title = "Non-Nutritive vs. No-Feed Visits",
    subtitle = "Each point represents one animal",
    x = "Number of Non-Nutritive Visits\n(feed available but not consumed)",
    y = "Number of No-Feed Visits\n(empty bin visits)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 10, color = "gray40")
  )
```

## 8. Temporal Patterns Across Days

Understanding how these patterns change over time can reveal important trends:

```{r temporal-analysis}
# Combine all days for temporal analysis
all_non_nutritive <- do.call(rbind, lapply(names(non_nutritive), function(date) {
  non_nutritive[[date]] |>
    dplyr::mutate(date = date)
}))

all_no_feed <- do.call(rbind, lapply(names(no_feed), function(date) {
  no_feed[[date]] |>
    dplyr::mutate(date = date)
}))

# Calculate daily averages
daily_summary <- all_non_nutritive |>
  dplyr::group_by(date) |>
  dplyr::summarize(
    avg_non_nutritive = mean(number_of_non_nutritive_visits),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    all_no_feed |>
      dplyr::group_by(date) |>
      dplyr::summarize(
        avg_no_feed = mean(number_of_visits_when_no_feed),
        .groups = "drop"
      ),
    by = "date"
  )

cat("Daily averages across all animals:\n")
print(daily_summary)
```

### Identify Most Selective Animals

```{r selective-animals}
# Calculate average non-nutritive visits per animal across all days
animal_selectivity <- all_non_nutritive |>
  dplyr::group_by(cow) |>
  dplyr::summarize(
    avg_non_nutritive_visits = mean(number_of_non_nutritive_visits),
    days_observed = dplyr::n(),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(avg_non_nutritive_visits))

cat("🏆 TOP 5 MOST SELECTIVE ANIMALS (highest non-nutritive visits):\n")
head(animal_selectivity, 5)

cat("\n🍽️ TOP 5 LEAST SELECTIVE ANIMALS (lowest non-nutritive visits):\n")
tail(animal_selectivity, 5)
```

## 9. Practical Interpretation

### High Non-Nutritive Visits May Indicate:

- **Feed quality issues**: Animals rejecting poor-quality or unpalatable feed
- **Feed selectivity**: Animals preferring certain feed types or freshness
- **Social displacement**: Animals being pushed away before consuming adequate amounts
- **Health concerns**: Sick animals showing reduced appetite despite checking bins

### High No-Feed Visits May Indicate:

- **Inadequate feed supply**: Bins frequently empty before all animals can eat
- **Intense feeding competition**: Dominant animals consuming available feed quickly
- **Poor feeding management**: Feed delivery timing not matching animal behavior patterns
- **Exploration behavior**: Animals actively searching for feeding opportunities

## 10. Summary

This tutorial demonstrated analysis of non-nutritive and no-feed visit patterns:

✅ **Non-nutritive visits**: Identified when animals visit bins with feed but don't eat

✅ **No-feed visits**: Identified when animals visit empty bins

✅ **Individual patterns**: Analyzed which animals show highest selectivity

✅ **Temporal trends**: Examined how patterns change across days

These metrics provide valuable insights into feeding selectivity, resource competition, and potential welfare issues.

## 11. Code Cheatsheet

```{r cheatsheet, eval=FALSE}
#' Copy and modify these code blocks for your own analysis!

# ---- SETUP: Global Variables (REQUIRED FIRST!) ----
library(moo4feed)
library(dplyr)
library(ggplot2)

# Set up your column names and timezone (modify these!)
set_global_cols(
  id_col = "cow",           # Your animal ID column
  start_col = "start",      # Visit start time column
  end_col = "end",          # Visit end time column
  bin_col = "bin",          # Bin/feeder ID column
  intake_col = "intake",    # Feed intake amount column
  dur_col = "duration",     # Visit duration column
  tz = "America/Vancouver"  # Your timezone
)

# ---- STEP 1: Load Your Data ----
# Use your own cleaned data from previous tutorials:
# clean_feed <- your_cleaned_feed_data

# ---- STEP 2: Create QC Configuration ----
my_qc_config <- qc_config(
  calibration_error = 0.5    # Equipment measurement threshold (kg)
)

# ---- STEP 3: Calculate Non-Nutritive Visits ----
# Visits when feed is available but animal doesn't eat
non_nutritive <- calculate_non_nutritive_visits(
  data = your_clean_feed,          # Your cleaned feed data
  cfg = my_qc_config,              # QC configuration
  id_col = id_col2(),              # Animal ID column (default from global vars)
  intake_col = intake_col2(),      # Intake column (default from global vars)
  start_weight_col = start_weight_col2()  # Start weight column (default from global vars)
)

# ---- STEP 4: Calculate No-Feed Visits ----
# Visits when bin is empty (no feed available)
no_feed <- calculate_no_feed_visits(
  data = your_clean_feed,          # Your cleaned feed data
  cfg = my_qc_config,              # QC configuration
  id_col = id_col2(),              # Animal ID column (default from global vars)
  intake_col = intake_col2(),      # Intake column (default from global vars)
  start_weight_col = start_weight_col2()  # Start weight column (default from global vars)
)

# ---- STEP 5: Combine and Analyze Results ----
# For a specific day
day1_combined <- non_nutritive[[1]] |>
  dplyr::full_join(no_feed[[1]], by = "cow") |>
  dplyr::mutate(
    number_of_non_nutritive_visits = tidyr::replace_na(number_of_non_nutritive_visits, 0),
    number_of_visits_when_no_feed = tidyr::replace_na(number_of_visits_when_no_feed, 0)
  )

# Identify most selective animals
most_selective <- day1_combined |>
  dplyr::arrange(dplyr::desc(number_of_non_nutritive_visits)) |>
  head(10)

# Animals checking empty bins most often
empty_bin_checkers <- day1_combined |>
  dplyr::arrange(dplyr::desc(number_of_visits_when_no_feed)) |>
  head(10)

# ---- STEP 6: Temporal Analysis Across Days ----
# Combine all days
all_non_nutritive <- do.call(rbind, lapply(names(non_nutritive), function(date) {
  non_nutritive[[date]] |>
    dplyr::mutate(date = date)
}))

all_no_feed <- do.call(rbind, lapply(names(no_feed), function(date) {
  no_feed[[date]] |>
    dplyr::mutate(date = date)
}))

# Calculate daily averages
daily_summary <- all_non_nutritive |>
  dplyr::group_by(date) |>
  dplyr::summarize(
    avg_non_nutritive = mean(number_of_non_nutritive_visits),
    .groups = "drop"
  ) |>
  dplyr::left_join(
    all_no_feed |>
      dplyr::group_by(date) |>
      dplyr::summarize(
        avg_no_feed = mean(number_of_visits_when_no_feed),
        .groups = "drop"
      ),
    by = "date"
  )

# Calculate per-animal averages across all days
animal_selectivity <- all_non_nutritive |>
  dplyr::group_by(cow) |>
  dplyr::summarize(
    avg_non_nutritive_visits = mean(number_of_non_nutritive_visits),
    days_observed = dplyr::n(),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(avg_non_nutritive_visits))

# ---- STEP 7: Visualization ----
# Scatter plot of visit types
ggplot(day1_combined, aes(x = number_of_non_nutritive_visits,
                          y = number_of_visits_when_no_feed)) +
  geom_point(alpha = 0.6, size = 3, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "coral", linetype = "dashed") +
  labs(
    title = "Non-Nutritive vs. No-Feed Visits",
    x = "Non-Nutritive Visits (feed available, not consumed)",
    y = "No-Feed Visits (empty bin)"
  ) +
  theme_minimal()
```

---

**🦦 Ollie's Final Words**: *"Remember, non-nutritive visits aren't always bad! They're natural exploration behavior. However, sudden increases in non-nutritive visits across many animals might signal feed quality issues or health concerns worth investigating."*

**Next Steps**: Use these metrics alongside other behavioral measures (from previous tutorials) to build a comprehensive picture of animal welfare and feeding patterns! 🍽️✨
