---
title: "4. Replacement Detection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4. Replacement Detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(moo4feed)
library(ggplot2)
library(dplyr)
```

## 1. 🚨 Important: Set Your Global Variables First!

Before analyzing replacement behavior, configure global variables to match your data structure:

```{r global-setup}
# Configure global variables for your data structure
set_global_cols(
  id_col = "cow",           # Animal ID column name
  start_col = "start",      # Visit start time column  
  end_col = "end",          # Visit end time column
  bin_col = "bin",          # Bin/feeder ID column
  intake_col = "intake",    # Feed intake amount column
  dur_col = "duration",     # Visit duration column
  tz = "America/Vancouver"  # Your timezone
)

# Verify configuration
cat("✅ Global variables configured:\n")
cat("ID column:", id_col2(), "\n")
cat("Start time column:", start_col2(), "\n") 
cat("End time column:", end_col2(), "\n")
cat("Bin column:", bin_col2(), "\n")
cat("Intake column:", intake_col2(), "\n")
cat("Duration column:", duration_col2(), "\n")
cat("Timezone:", tz2(), "\n")
```

## 2. Introduction to Replacement Detection

🦦 Ollie the Otter explains: "Replacement events occur when one animal pushes and displaces another animal that is actively feedingfrom a feeding station within a short time window. These social interactions reveal important information about herd dynamics and dominance hierarchies."

Understanding replacement behavior helps researchers and farm managers:
- **Monitor social stress**: High replacement rates may indicate social tension
- **Identify dominant animals**: Animals that frequently replace others
- **Affective state**: Animals frequently get replaced may experience longitudinal stress

### What We'll Learn

This tutorial demonstrates how to: **Detect replacement events** using validated time thresholds

## 3. Prerequisites

This tutorial assumes completion of previous data processing steps in **Tutorial 1: Data Cleaning** 


## 4. Data Preparation

```{r data-prep}
# Load cleaned example data
data(clean_feed)
data(clean_water)

# If you're using your own data from previous tutorials, use this instead:
# clean_feed <- your_cleaned_feed_data     # From your cleaning results
# clean_water <- your_cleaned_water_data   # From your cleaning results

# Quick peek at our data structure
cat("Feed data structure:\n")
head(clean_feed[[1]], 3)  # First day, first 3 rows

cat("\nTotal days of feed data:", length(clean_feed), "\n")
```

## 5. Understanding Replacement Events

A replacement event is defined as when one animal (actor) takes over a feeding station from another animal (reactor) within a short time threshold. The default threshold of 26 seconds is based on validated research.

### Key Components

- **Reactor cow**: The cow that was replaced (had to leave the bin)
- **Actor cow**: The cow that initiated the replacement (took over the bin)  
- **Time threshold**: Maximum gap between when one cow leaves and another arrives (default: 26 seconds)
- **Validation**: Events are verified by checking if the actor cow has an "alibi" (was feeding elsewhere when the reactor left)

### Detecting Replacement Events

```{r replacement-analysis}
# Process replacement events for all days
replacements <- record_replacement_days(
  comb = clean_feed,           # Our cleaned feed data
  cfg = qc_config(replacement_threshold = 26)      # Time gap (seconds) to classify replacement behavior
)

# Examine the first few replacement events
head(replacements[[1]])

# Summary of replacement events
cat("Replacement events per day:\n")
sapply(replacements, nrow)
```

Each replacement event contains:
- **`reactor_cow`**: ID of the cow that was replaced
- **`actor_cow`**: ID of the cow that initiated the replacement
- **`bin`**: Feeding/drinking station where the replacement occurred
- **`time`**: Timestamp when the replacement happened
- **`date`**: Date of the event
- **`bout_interval`**: Time gap between when the reactor left and actor entered the bin

## 6. Analyzing Replacement Patterns

### Most Active cows

```{r replacement-patterns}
# Combine all days for analysis
all_replacements <- do.call(rbind, replacements)

# cows that most frequently replace others (actors)
top_actors <- all_replacements |>
  dplyr::count(actor_cow, sort = TRUE, name = "times_replaced_others") |>
  head(5)

cat("Top 10 cows that most frequently displace others:\n")
print(top_actors)

# cows that are most frequently replaced (reactors)
top_reactors <- all_replacements |>
  dplyr::count(reactor_cow, sort = TRUE, name = "times_replaced") |>
  head(5)

cat("\nTop 10 cows that are most frequently displaced:\n")
print(top_reactors)
```

### Replacement Timing Patterns

```{r timing-patterns}
# Analyze replacement timing throughout the day
all_replacements$hour <- lubridate::hour(all_replacements$time)

hourly_replacements <- all_replacements |>
  dplyr::count(hour, name = "replacement_count")

cat("Replacement events by hour of day:\n")
print(hourly_replacements)
```

## 7. Summary

This tutorial demonstrated replacement detection for understanding social dynamics:

✅ **Event detection**: Identified and validated replacement events using time thresholds
✅ **Pattern analysis**: Analyzed which cows are most active in displacement behavior
✅ **Timing insights**: Examined when most replacement events occur throughout the day

## 8. Code Cheatsheet

```{r cheatsheet, eval=FALSE}
#' Copy and modify these code blocks for your own analysis!

# ---- SETUP: Global Variables (REQUIRED FIRST!) ----
library(moo4feed)
library(dplyr)

# Set up your column names and timezone (modify these!)
set_global_cols(
  id_col = "cow",           # Your animal ID column
  start_col = "start",      # Visit start time column  
  end_col = "end",          # Visit end time column
  bin_col = "bin",          # Bin/feeder ID column
  intake_col = "intake",    # Feed intake amount column
  dur_col = "duration",     # Visit duration column
  tz = "America/Vancouver"  # Your timezone
)

# ---- STEP 1: Load Your Data ----
# Use your own cleaned data from previous tutorials:
# clean_feed <- your_cleaned_feed_data

# ---- STEP 2: Detect Replacement Events ----
# Process replacement events for all days
replacements <- record_replacement_days(
  comb = your_clean_feed,         # Your cleaned feed data
  cfg = qc_config(replacement_threshold = 26),     # Time gap (seconds) to classify replacement behavior
  id_col = id_col2(),             # Animal ID column (default from global vars)
  bin_col = bin_col2(),           # Bin/feeder ID column (default from global vars)
  start_col = start_col2(),       # Visit start time column (default from global vars)
  end_col = end_col2()            # Visit end time column (default from global vars)
)

# Check results
cat("Total replacement events found:\n")
sapply(replacements, nrow)

# Look at first few events from first day
head(replacements[[1]])

# ---- STEP 3: Analyze Replacement Patterns ----
# Combine all days for analysis
all_replacements <- do.call(rbind, replacements)

# Find most active displacing animals
top_actors <- all_replacements |>
  dplyr::count(actor_cow, sort = TRUE, name = "times_displaced_others") |>
  head(10)

# Find most frequently displaced animals  
top_reactors <- all_replacements |>
  dplyr::count(reactor_cow, sort = TRUE, name = "times_displaced") |>
  head(10)

# Analyze timing patterns
all_replacements$hour <- lubridate::hour(all_replacements$time)
hourly_replacements <- all_replacements |>
  dplyr::count(hour, name = "replacement_count")

# ---- STEP 4: Quick Analysis ----
# Check replacement results
cat("Replacement events per day:\n")
sapply(replacements, nrow)

cat("Most active displacing animals:\n")
print(top_actors)

cat("Most frequently displaced animals:\n")
print(top_reactors)
``` 