---
title: "4. Replacement Detection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4. Replacement Detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(moo4feed)
library(ggplot2)
library(dplyr)
```

## 🚨 Important: Set Your Global Variables First!

*"Before we investigate social dynamics," Ollie says while putting on his detective hat, "make sure you've configured your global variables to match your data structure. This makes all the functions work smoothly!"*

If you haven't done this yet, run the setup from **Tutorial 1: Data Cleaning**:

```{r global-setup}
# 🎯 IMPORTANT: Set these to match YOUR data structure!
# Copy and modify these lines to match your column names and settings:

set_global_cols(
  id_col = "cow",           # Animal ID column name
  start_col = "start",      # Visit start time column  
  end_col = "end",          # Visit end time column
  bin_col = "bin",          # Bin/feeder ID column
  intake_col = "intake",    # Feed intake amount column
  dur_col = "duration",     # Visit duration column
  tz = "America/Vancouver"  # Your timezone
)

# Check that everything is set correctly:
cat("✅ Global variables set:\n")
cat("ID column:", id_col2(), "\n")
cat("Start time column:", start_col2(), "\n") 
cat("End time column:", end_col2(), "\n")
cat("Bin column:", bin_col2(), "\n")
cat("Intake column:", intake_col2(), "\n")
cat("Duration column:", duration_col2(), "\n")
cat("Timezone:", tz2(), "\n")
```

*"Perfect! Now all our functions will automatically use the right column names for your data!"* 🦦

## Ollie's Social Detective Mission 🕵️‍♂️

*"Welcome back, my fabulous data detectives!"* Ollie adjusts his magnifying glass and detective hat. *"After exploring meal clustering and bin visit patterns, I discovered something fascinating about animal social behavior! You know how sometimes you're enjoying your favorite spot at a café, and someone comes and takes the table right next to you, making you feel like you need to move? Animals do this too!"*

*"We call these 'replacement events' - when one animal displaces another from a feeding or drinking station. This behavior reveals so much about social hierarchies, stress levels, and individual personalities!"*

### What We'll Discover Today

- **Social hierarchies**: Who are the bold displacing animals vs. the ones that get displaced?
- **Social interactions**: How do replacement events reveal herd dynamics?
- **Behavioral insights**: What can replacement patterns tell us about animal welfare?

Let's solve this social mystery! 🔍

## Prerequisites: Clean Data First! 🧹

⚠️ **Important:** This tutorial assumes you have already completed:
- **Tutorial 1: Data Cleaning** 
- **Tutorial 2: Meal Clustering**
- **Tutorial 3: Bin Visit Analysis**

```{r data-prep}
# Load the example cleaned data from the package
data(clean_feed)
data(clean_water)

# If you're using your own data from previous tutorials, use this instead:
# clean_feed <- your_cleaned_feed_data     # From your cleaning results
# clean_water <- your_cleaned_water_data   # From your cleaning results

# Quick peek at our data structure
cat("Feed data structure:\n")
head(clean_feed[[1]], 3)  # First day, first 3 rows

cat("\nTotal days of feed data:", length(clean_feed), "\n")
```

## Understanding Replacement Events 🏆

*"Here's where the detective work gets exciting!"* Ollie grins while polishing his magnifying glass. *"A replacement event happens when one animal takes over a feeding station from another animal within a short time window. We need to be smart about this - we don't want to count coincidental visits as social events!"*

Dairy animals establish complex social hierarchies. One key social behavior is when one animal replaces another at feeding or drinking stations.

### Detecting Replacement Events

Now, let's use the `record_replacement_days()` function to detect replacement events across all days in our dataset:

- The `replacement_threshold` parameter is critical - if an animal is feeding at a bin, and the next animal that enters the same bin within 26 seconds, it is considered a replacement event. This threshold is validated by previous studies.

```{r replacement-analysis}
# Process feed replacement events for all days
replacements <- record_replacement_days(
  comb = clean_feed,           # Our cleaned feed data
  cfg = qc_config(replacement_threshold = 26)      # Time gap (seconds) to classify replacement behavior
)

# Look at the first few replacement events for the first day
head(replacements[[1]])
```

*"Isn't this fascinating?"* Ollie claps excitedly. *"Each row represents a social interaction where:*

- **`reactor_cow`**: The animal that was replaced (had to leave the bin)
- **`actor_cow`**: The animal that initiated the replacement (took over the bin)  
- **`bin`**: Which feeding station this drama happened at
- **`time`**: When this social event occurred
- **`date`**: Which day this happened
- **`bout_interval`**: The exact time gap between when one animal left and the next arrived

*The function even validates each event by checking if the displacing animal has an 'alibi' - making sure they weren't already feeding elsewhere when the previous animal left!"*

## What Replacement Events Tell Us 📊

*"These social interactions reveal so much about our animals!"* Ollie explains while examining the data.

Patterns in replacement events can reveal:

- Days with high social tension in the herd
- Changes in social dynamics over time
- Effects of management changes on social behavior
- Potential welfare issues when replacement events spike

Importantly, the function validates each replacement event by checking if the actor animal has an "alibi" - if the animal was recorded at another bin when the previous animal left, the event is excluded. This ensures we're capturing true social replacement events rather than coincidental timings.

## 🎉 Ollie's Social Detective Summary!

*"What an incredible investigation!"* Ollie tips his detective hat. *"We've uncovered so many secrets about animal social dynamics! Now we can clearly see:*

- *How to detect and validate replacement events*
- *What these social interactions tell us about herd dynamics*
- *How to identify patterns in social behavior*

*"This information is incredibly valuable for:*
- *Understanding animal welfare and stress*
- *Identifying management strategies to reduce social pressure*
- *Recognizing individual animal personalities and needs*
- *Monitoring changes in herd dynamics over time*

*Remember, replacement events aren't necessarily 'bad' - they're natural social behaviors! But understanding these patterns helps us manage our animals better."*

**What we've accomplished together:**
- ✅ Learned to detect and validate replacement events
- ✅ Understood what social interactions reveal about herd dynamics
- ✅ Discovered how to analyze social behavior patterns

*"The best part is you can now use these techniques to understand social dynamics in your own herds!"* 🦦

## 🦦 Replacement Detection Cheat Sheet

*"Here's my complete reference guide for replacement detection and social analysis! Copy and paste these code blocks to work with your own data."*

```r
#' ============================================================================
#' 🦦 OLLIE'S REPLACEMENT DETECTION CHEAT SHEET
#' ============================================================================
#' Copy and modify these code blocks for your own analysis!

# ---- SETUP: Global Variables (REQUIRED FIRST!) ----
library(moo4feed)
library(dplyr)

# Set up your column names and timezone (modify these!)
set_global_cols(
  id_col = "cow",           # Your animal ID column
  start_col = "start",      # Visit start time column  
  end_col = "end",          # Visit end time column
  bin_col = "bin",          # Bin/feeder ID column
  intake_col = "intake",    # Feed intake amount column
  dur_col = "duration",     # Visit duration column
  tz = "America/Vancouver"  # Your timezone
)

# ---- STEP 1: Load Your Data ----
# Use your own cleaned data from previous tutorials:
# clean_feed <- your_cleaned_feed_data

# ---- STEP 2: Detect Replacement Events ----
# Process replacement events for all days
replacements <- record_replacement_days(
  comb = your_clean_feed,         # Your cleaned feed data
  cfg = qc_config(replacement_threshold = 26)      # Time gap (seconds) to classify replacement behavior
)

# Check results
cat("Total replacement events found:\n")
sapply(replacements, nrow)

# Look at first few events from first day
head(replacements[[1]])

# ---- STEP 3: Quick Analysis ----
# Check replacement results
cat("Replacement events per day:\n")
sapply(replacements, nrow)

#' ============================================================================
#' 🦦 That's it! You now understand social dynamics in your herd!
#' ============================================================================
``` 