---
title: "5. Synchronicity Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Synchronicity Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(moo4feed)
library(ggplot2)
library(dplyr)
```

## 1. 🚨 Important: Set Your Global Variables First!

Before analyzing synchronicity patterns, configure global variables to match your data structure:

```{r global-setup}
# Configure global variables for your data structure
set_global_cols(
  id_col = "cow",           # Animal ID column name
  start_col = "start",      # Visit start time column
  end_col = "end",          # Visit end time column
  bin_col = "bin",          # Bin/feeder ID column
  start_weight_col = "start_weight",  # Start weight column
  end_weight_col = "end_weight",      # End weight column
  tz = "America/Vancouver", # Your timezone
  bins_feed = 1:30,         # Feed bin IDs
  bins_wat = 1:5,           # Water bin IDs
  bin_layout = "1-2-3-4-5-6-101-102-7-8-9-10-11-12-13-14-15-16-17-18-103-104-19-20-21-22-23-24-25-26-27-28-29-30-105"
)

# Verify configuration
cat("✅ Global variables configured:\n")
cat("ID column:", id_col2(), "\n")
cat("Start time column:", start_col2(), "\n")
cat("End time column:", end_col2(), "\n")
cat("Bin column:", bin_col2(), "\n")
cat("Timezone:", tz2(), "\n")
cat("Bin layout:", bin_layout2(), "\n")
```

## 2. Introduction to Synchronicity Analysis

🦦 Ollie the Otter explains: "Synchronicity in animal behavior refers to animals performing the same activity at the same time. For feeding and drinking behavior, synchronicity can reveal social dynamics in the herd. Understanding synchronicity helps researchers identify social bonds and competition patterns."

### What Is Synchronicity?

Synchronicity analysis examines temporal overlap in feeding or drinking activities between animals. There are two main types:

> 1. **Pair-wise co-occurrence**: How much time do pairs of animals spend feeding or drinking simultaneously, regardless of location?
> 2. **Spatial neighbor proximity**: How much time do pairs spend at neighboring bins (spatially close)?

### Why Study Synchronicity?

- **Social structure**: Identify socially bonded pairs that tend to feed together
- **Competition**: Detect animals that avoid each other during feeding
- **Spatial preferences**: Understand how animals use feeding space
- **Welfare monitoring**: Changes in synchronicity may indicate social stress or health issues

## 3. Prerequisites

This tutorial assumes you have completed data cleaning from **Tutorial 1: Data Cleaning**. You need cleaned feeding or drinking data.

## 4. Data Preparation

```{r data-prep}
# Load cleaned example data
data(clean_feed)
data(clean_water)

# If using your own data from previous tutorials:
# clean_feed <- your_cleaned_feed_data
# clean_water <- your_cleaned_water_data

# Quick peek at data structure
cat("Feed data structure:\n")
head(clean_feed[[1]], 3)

cat("\nTotal days of feed data:", length(clean_feed), "\n")
cat("Total days of water data:", length(clean_water), "\n")
```

## 5. Creating Time-Based Activity Matrices

Before analyzing synchronicity, we need to transform visit-based data into time-based matrices. The `matrix_process()` function creates matrices where:

- **Rows** represent time points (by second or minute)
- **Columns** represent individual animals
- **Values** indicate activity status at each time point

### Understanding Matrix Types

`matrix_process()` creates three types of matrices:

> 1. **Animal activity matrix** (`synch_master_animal2`): Shows which animals are active at each time point (1 = active, 0 = inactive)
> 2. **Bin occupancy matrix** (`synch_master_bin2`): Shows which bin each animal is using at each time point (0 = not active)
> 3. **Feed weight matrix** (`synch_master_feed2`): For feed data only, shows feed weight at each bin over time

### Process Feed Data Matrices

```{r process-feed-matrices}
# Process feed data to create time-based matrices
# Resolution "sec" means one row per second (more detailed but larger)
# Resolution "min" means one row per minute (less detailed but smaller)
feed_matrices <- matrix_process(
  data_list = clean_feed,
  type = "feed",
  resolution = "sec",  # Try "min" for faster processing with less detail
  id_col = id_col2(),
  start_col = start_col2(),
  end_col = end_col2(),
  bin_col = bin_col2(),
  start_weight_col = start_weight_col2(),
  end_weight_col = end_weight_col2(),
  bins_feed = bins_feed2()
)

# Examine the structure
cat("Matrix components created:\n")
names(feed_matrices)

# Look at the animal activity matrix for the first day
cat("\nAnimal activity matrix structure (first 5 rows):\n")
head(feed_matrices$synch_master_animal2[[1]][, 1:5], 5)

cat("\nColumns in animal matrix:\n")
names(feed_matrices$synch_master_animal2[[1]])
```

### Process Water Data Matrices

```{r process-water-matrices}
# Process water data similarly
water_matrices <- matrix_process(
  data_list = clean_water,
  type = "drink",
  resolution = "sec",
  id_col = id_col2(),
  start_col = start_col2(),
  end_col = end_col2(),
  bin_col = bin_col2(),
  bins_wat = bins_wat2()
)

cat("Water matrix components created:\n")
names(water_matrices)
```

### Understanding the Output

Each matrix type provides different information:

- **Animal matrix**: Use this for pair-wise co-occurrence analysis (who feeds together?)
- **Bin matrix**: Use this for spatial neighbor analysis (who feeds near each other?)
- **Feed matrix**: Additional information about feed depletion patterns

The matrices are filtered to include only time points when at least one animal is active, reducing memory usage.

## 6. Pair-wise Co-Occurrence Analysis

Pair-wise analysis examines how much time each pair of animals spends feeding or drinking simultaneously, regardless of their spatial location.

### Analyze Feed Synchronicity

```{r pair-feed-analysis}
# Analyze pair-wise co-occurrence for feeding
pair_feed_results <- synch_pair_analysis(
  matrix_data = feed_matrices,
  type = "feed",
  resolution = "sec",
  id_col = id_col2()
)

# Examine the structure
cat("Pair analysis results:\n")
names(pair_feed_results)

# Look at bout counts for the first day
cat("\nBout matrix (first 5 animals, first day):\n")
bout_matrix_day1 <- pair_feed_results$bout[[1]]
print(bout_matrix_day1[1:5, 1:5])

# Look at total time together
cat("\nTotal time together matrix (seconds, first 5 animals, first day):\n")
time_matrix_day1 <- pair_feed_results$total_time[[1]]
print(time_matrix_day1[1:5, 1:5])

# Look at average duration per bout
cat("\nAverage duration per bout (seconds, first 5 animals, first day):\n")
avg_matrix_day1 <- pair_feed_results$avg_duration[[1]]
print(avg_matrix_day1[1:5, 1:5])
```

### Interpreting Pair Analysis Results

The results contain three matrices for each day:

> - **bout**: Number of distinct feeding bouts each pair spent together. A new bout starts when animals stop feeding together for more than 1 second (or 1 minute if using "min" resolution).
> - **total_time**: Total time (in seconds or minutes) each pair spent feeding simultaneously.
> - **avg_duration**: Average duration per bout for each pair.

**Important**: Only the upper triangle of each matrix is filled (values above the diagonal). The diagonal and lower triangle are 0 because:
- Diagonal: An animal doesn't form a pair with itself
- Lower triangle: Avoids redundancy (pair A-B is same as B-A)

### Finding Most Synchronized Pairs

```{r find-synchronized-pairs}
# Find pairs with the most time spent together (first day)
# Convert upper triangle to data frame for easier analysis
pair_df <- data.frame(
  animal1 = character(),
  animal2 = character(),
  bouts = numeric(),
  total_time = numeric(),
  avg_duration = numeric()
)

# Extract upper triangle values
animal_ids <- rownames(time_matrix_day1)
for (i in 1:(length(animal_ids) - 1)) {
  for (j in (i + 1):length(animal_ids)) {
    if (time_matrix_day1[i, j] > 0) {  # Only include pairs with activity
      pair_df <- rbind(pair_df, data.frame(
        animal1 = animal_ids[i],
        animal2 = animal_ids[j],
        bouts = bout_matrix_day1[i, j],
        total_time = time_matrix_day1[i, j],
        avg_duration = avg_matrix_day1[i, j]
      ))
    }
  }
}

# Sort by total time together
pair_df <- pair_df[order(-pair_df$total_time), ]

cat("Top 10 most synchronized feeding pairs (first day):\n")
print(head(pair_df, 10))
```

### Analyze Water Drinking Synchronicity

```{r pair-water-analysis}
# Analyze pair-wise co-occurrence for drinking
pair_water_results <- synch_pair_analysis(
  matrix_data = water_matrices,
  type = "drink",
  resolution = "sec",
  id_col = id_col2()
)

# Look at total time together for water (first day)
water_time_matrix <- pair_water_results$total_time[[1]]
cat("Water drinking synchronicity (first 5 animals, first day):\n")
print(water_time_matrix[1:5, 1:5])
```

## 7. Spatial Neighbor Proximity Analysis

Neighbor analysis examines how much time animals spend at spatially adjacent bins. This requires the bin layout configuration you set earlier.

### Understanding Bin Layout

The bin layout string defines the spatial arrangement of bins:
- Rows are separated by `\n` (newline)
- Bins within a row are separated by `-`
- Only bins in the same row are considered neighbors (left/right adjacency)

Example: `"1-2-3-4-5-6-101-102-7-8-9-10"` means bins 1, 2, 3, 4, 5, 6, 101, 102, 7, 8, 9, 10 are in a single row.
- Bin 1 is neighbor to bin 2 only
- Bin 5 is neighbor to bins 4 and 6
- Bin 101 is neighbor to bins 6 and 102

```{r show-bin-layout}
cat("Current bin layout:\n")
cat(bin_layout2(), "\n")
```

### Analyze Feed Neighbor Synchronicity

```{r neighbor-feed-analysis}
# Analyze spatial neighbor patterns for feeding
neighbor_feed_results <- synch_neighbor_analysis(
  matrix_data = feed_matrices,
  bin_layout = bin_layout2(),
  type = "feed",
  resolution = "sec",
  id_col = id_col2()
)

# Examine the structure
cat("Neighbor analysis results:\n")
names(neighbor_feed_results)

# Look at bout counts for neighbors (first day)
cat("\nNeighbor bout matrix (first 5 animals, first day):\n")
neighbor_bout_matrix <- neighbor_feed_results$bout[[1]]
print(neighbor_bout_matrix[1:5, 1:5])

# Look at total time as neighbors
cat("\nTotal time as neighbors (seconds, first 5 animals, first day):\n")
neighbor_time_matrix <- neighbor_feed_results$total_time[[1]]
print(neighbor_time_matrix[1:5, 1:5])
```

### Comparing Co-Occurrence vs. Neighbor Patterns

```{r compare-patterns}
# Compare the same pair in both analyses
example_animal1 <- rownames(time_matrix_day1)[1]
example_animal2 <- rownames(time_matrix_day1)[2]

cat("Comparison for animals", example_animal1, "and", example_animal2, "(first day):\n\n")

cat("Co-occurrence (feeding at same time, any bins):\n")
cat("  Bouts:", bout_matrix_day1[example_animal1, example_animal2], "\n")
cat("  Total time:", time_matrix_day1[example_animal1, example_animal2], "seconds\n")
cat("  Avg duration:", round(avg_matrix_day1[example_animal1, example_animal2], 1), "seconds\n\n")

cat("Neighbor proximity (feeding at adjacent bins):\n")
cat("  Bouts:", neighbor_bout_matrix[example_animal1, example_animal2], "\n")
cat("  Total time:", neighbor_time_matrix[example_animal1, example_animal2], "seconds\n")
cat("  Avg duration:", round(neighbor_feed_results$avg_duration[[1]][example_animal1, example_animal2], 1), "seconds\n")
```

### Finding Pairs That Prefer Neighboring Bins

```{r neighbor-preference}
# Find pairs with high neighbor time relative to total co-occurrence time
neighbor_df <- data.frame(
  animal1 = character(),
  animal2 = character(),
  cooccurrence_time = numeric(),
  neighbor_time = numeric(),
  neighbor_ratio = numeric()
)

# Extract pairs with both types of activity
for (i in 1:(length(animal_ids) - 1)) {
  for (j in (i + 1):length(animal_ids)) {
    co_time <- time_matrix_day1[i, j]
    nb_time <- neighbor_time_matrix[i, j]

    if (co_time > 0) {  # Only pairs with some co-occurrence
      neighbor_df <- rbind(neighbor_df, data.frame(
        animal1 = animal_ids[i],
        animal2 = animal_ids[j],
        cooccurrence_time = co_time,
        neighbor_time = nb_time,
        neighbor_ratio = nb_time / co_time  # Proportion of time as neighbors
      ))
    }
  }
}

# Sort by neighbor ratio
neighbor_df <- neighbor_df[order(-neighbor_df$neighbor_ratio), ]

cat("Top 10 pairs with highest neighbor preference (first day):\n")
cat("(Higher ratio = more time at neighboring bins when feeding together)\n")
print(head(neighbor_df, 10))
```

## 8. Analyzing Multiple Days

Both analysis functions handle multi-day data automatically. You can analyze patterns across days.

```{r multi-day-analysis}
# Compare synchronicity across days
cat("Total feeding co-occurrence time by day:\n")
for (i in seq_along(pair_feed_results$total_time)) {
  day_name <- names(pair_feed_results$total_time)[i]
  day_matrix <- pair_feed_results$total_time[[i]]
  # Sum upper triangle only to avoid double counting
  total_time <- sum(day_matrix[upper.tri(day_matrix)])
  cat(day_name, ":", total_time, "seconds\n")
}

cat("\nTotal neighbor time by day:\n")
for (i in seq_along(neighbor_feed_results$total_time)) {
  day_name <- names(neighbor_feed_results$total_time)[i]
  day_matrix <- neighbor_feed_results$total_time[[i]]
  total_time <- sum(day_matrix[upper.tri(day_matrix)])
  cat(day_name, ":", total_time, "seconds\n")
}
```

## 9. Practical Applications

### Identifying Social Bonds

Animals that consistently spend high amounts of time feeding together may have social bonds:

```{r social-bonds}
# Calculate average co-occurrence time across all days
if (length(pair_feed_results$total_time) > 1) {
  # Sum across days for each pair
  avg_time_matrix <- pair_feed_results$total_time[[1]] * 0  # Initialize with zeros

  for (day_matrix in pair_feed_results$total_time) {
    avg_time_matrix <- avg_time_matrix + day_matrix
  }

  # Divide by number of days
  avg_time_matrix <- avg_time_matrix / length(pair_feed_results$total_time)

  # Find consistently synchronized pairs
  cat("Pairs with consistently high synchronicity (average > 100 seconds/day):\n")
  for (i in 1:(nrow(avg_time_matrix) - 1)) {
    for (j in (i + 1):ncol(avg_time_matrix)) {
      if (avg_time_matrix[i, j] > 100) {
        cat(rownames(avg_time_matrix)[i], "-",
            colnames(avg_time_matrix)[j], ":",
            round(avg_time_matrix[i, j], 1), "seconds/day\n")
      }
    }
  }
}
```

### Detecting Competition

Low synchronicity despite being in the same herd might indicate avoidance or competition:

```{r detect-competition}
# Find pairs with very low co-occurrence (potential avoidance)
cat("Pairs with minimal co-occurrence (< 10 seconds total across all days):\n")
low_sync_count <- 0

if (length(pair_feed_results$total_time) > 1) {
  total_time_matrix <- pair_feed_results$total_time[[1]] * 0

  for (day_matrix in pair_feed_results$total_time) {
    total_time_matrix <- total_time_matrix + day_matrix
  }

  for (i in 1:(nrow(total_time_matrix) - 1)) {
    for (j in (i + 1):ncol(total_time_matrix)) {
      if (total_time_matrix[i, j] < 10 && total_time_matrix[i, j] >= 0) {
        low_sync_count <- low_sync_count + 1
      }
    }
  }

  cat("Total pairs with low synchronicity:", low_sync_count, "\n")
}
```

## 10. Tips and Best Practices

> **Resolution Choice**:
> - Use `resolution = "sec"` for detailed analysis, but requires more memory
> - Use `resolution = "min"` for faster processing with larger datasets
> - Resolution affects bout detection: "sec" detects gaps > 1 second, "min" detects gaps > 60 seconds

> **Bin Layout Configuration**:
> - Ensure your bin layout matches the physical arrangement in your barn
> - Only bins in the same row are considered neighbors
> - Double-check bin numbers match those in your data

> **Memory Considerations**:
> - Time-based matrices can be large for long time periods at second resolution
> - Consider processing one day at a time for very large datasets
> - Use minute resolution for preliminary exploration

> **Interpretation**:
> - High co-occurrence doesn't always mean social bonding (could be just feeding pattern overlap)
> - Compare against null models or randomized data for statistical significance
> - Consider both bout count and duration when interpreting synchronicity

## 11. Summary

This tutorial demonstrated synchronicity analysis for understanding social dynamics in feeding and drinking behavior:

✅ **Matrix creation**: Transformed visit data into time-based activity matrices

✅ **Pair-wise analysis**: Quantified how much time animal pairs spend feeding or drinking together

✅ **Neighbor analysis**: Analyzed spatial proximity patterns at adjacent bins

✅ **Multi-day patterns**: Compared synchronicity across multiple days

✅ **Practical applications**: Identified potential social bonds and competition patterns

## 12. Code Cheatsheet

```{r cheatsheet, eval=FALSE}
# Copy and modify these code blocks for your own analysis!

# ---- SETUP: Global Variables (REQUIRED FIRST!) ----
library(moo4feed)
library(dplyr)

# Set up your column names and bin configuration
set_global_cols(
  id_col = "cow",           # Your animal ID column
  start_col = "start",      # Visit start time column
  end_col = "end",          # Visit end time column
  bin_col = "bin",          # Bin/feeder ID column
  start_weight_col = "start_weight",  # Start weight column (feed only)
  end_weight_col = "end_weight",      # End weight column (feed only)
  tz = "America/Vancouver", # Your timezone
  bins_feed = 1:30,         # Your feed bin IDs
  bins_wat = 1:5,           # Your water bin IDs
  # Define your barn's bin layout (rows separated by \n, bins by -)
  bin_layout = "1-2-3-4-5-6-101-102-7-8-9-10-11-12"
)

# ---- STEP 1: Create Time-Based Matrices ----
# For feed data
feed_matrices <- matrix_process(
  data_list = your_clean_feed,   # Your cleaned feed data
  type = "feed",
  resolution = "sec",              # Try: "min" for faster processing
  id_col = id_col2(),
  start_col = start_col2(),
  end_col = end_col2(),
  bin_col = bin_col2(),
  start_weight_col = start_weight_col2(),
  end_weight_col = end_weight_col2(),
  bins_feed = bins_feed2()
)

# For water data
water_matrices <- matrix_process(
  data_list = your_clean_water,   # Your cleaned water data
  type = "drink",
  resolution = "sec",
  id_col = id_col2(),
  start_col = start_col2(),
  end_col = end_col2(),
  bin_col = bin_col2(),
  bins_wat = bins_wat2()
)

# ---- STEP 2: Pair-wise Co-Occurrence Analysis ----
# Analyze feeding synchronicity
pair_feed_results <- synch_pair_analysis(
  matrix_data = feed_matrices,
  type = "feed",
  resolution = "sec",              # Must match matrix_process resolution
  id_col = id_col2()
)

# Analyze drinking synchronicity
pair_water_results <- synch_pair_analysis(
  matrix_data = water_matrices,
  type = "drink",
  resolution = "sec",
  id_col = id_col2()
)

# Access results (one matrix per day if multi-day input)
bout_counts <- pair_feed_results$bout           # Number of bouts together
total_times <- pair_feed_results$total_time    # Total time together
avg_durations <- pair_feed_results$avg_duration # Average bout duration

# For single day: bout_counts is a matrix
# For multi-day: bout_counts[[1]] is first day's matrix

# ---- STEP 3: Spatial Neighbor Analysis ----
# Analyze neighbor patterns for feeding
neighbor_feed_results <- synch_neighbor_analysis(
  matrix_data = feed_matrices,
  bin_layout = bin_layout2(),      # Use your configured layout
  type = "feed",
  resolution = "sec",
  id_col = id_col2()
)

# Analyze neighbor patterns for drinking
neighbor_water_results <- synch_neighbor_analysis(
  matrix_data = water_matrices,
  bin_layout = bin_layout2(),
  type = "drink",
  resolution = "sec",
  id_col = id_col2()
)

# Access neighbor results
neighbor_bouts <- neighbor_feed_results$bout
neighbor_times <- neighbor_feed_results$total_time
neighbor_avg <- neighbor_feed_results$avg_duration

# ---- STEP 4: Extract and Analyze Specific Pairs ----
# Get results for first day
day1_matrix <- pair_feed_results$total_time[[1]]

# Find value for a specific pair (e.g., animals "5120" and "6084")
pair_time <- day1_matrix["5120", "6084"]  # Time they spent together

# Find all pairs with high synchronicity
threshold <- 100  # seconds
high_sync_pairs <- which(day1_matrix > threshold, arr.ind = TRUE)
high_sync_pairs <- high_sync_pairs[high_sync_pairs[,1] < high_sync_pairs[,2], ]  # Upper triangle only

# ---- STEP 5: Compare Across Days ----
# Sum synchronicity across all days for each pair
if (length(pair_feed_results$total_time) > 1) {
  total_matrix <- pair_feed_results$total_time[[1]] * 0  # Initialize

  for (day_matrix in pair_feed_results$total_time) {
    total_matrix <- total_matrix + day_matrix
  }

  # Average across days
  avg_matrix <- total_matrix / length(pair_feed_results$total_time)
}

# ---- BONUS: Quick Summary Statistics ----
# Total synchronicity in the herd (first day)
day1_total <- sum(day1_matrix[upper.tri(day1_matrix)])
cat("Total co-occurrence time:", day1_total, "seconds\n")

# Number of active pairs (first day)
n_active_pairs <- sum(day1_matrix[upper.tri(day1_matrix)] > 0)
cat("Number of pairs with co-occurrence:", n_active_pairs, "\n")

# Average synchronicity per active pair
avg_sync <- day1_total / n_active_pairs
cat("Average time per active pair:", round(avg_sync, 1), "seconds\n")
```

---

**🦦 Ollie's Final Words**: *"Synchronicity analysis opens a window into the social world of your animals! Remember that feeding together doesn't always mean friendship—it could also be feeding pattern overlap. Combine synchronicity data with other behavioral measures for a complete picture of social dynamics!"*

**Next Steps**: Use synchronicity results to identify animals for detailed behavioral observation, detect changes in social structure over time, or validate social network models.
