---
title: "5. Synchronicity Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{5. Synchronicity Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(moo4feed)
library(ggplot2)
library(dplyr)
```

## 1. 🚨 Important: Set Your Global Variables First

Before analyzing synchronicity patterns, configure global variables to match your data structure:

```{r global-setup}
# Configure global variables for your data structure
set_global_cols(
  # Time zone
  tz = "America/Vancouver",        # Your timezone
  
  # Column names in your cleaned data
  id_col = "cow",
  trans_col = "transponder",
  start_col = "start",
  end_col = "end",
  bin_col = "bin",
  dur_col = "duration",
  intake_col = "intake",
  start_weight_col = "start_weight",
  end_weight_col = "end_weight",
  
  # Bin settings
  bins_feed = 1:30,                # All feed bin IDs in your barn
  bins_wat = 1:5,                  # All water bin IDs in your barn
  bin_offset = 100,                # Add this to water bin IDs to avoid conflicts
  
  # Physical bin layout for spatial neighbor analysis
  # Rows separated by "\n", bins within rows separated by "-"
  # Only bins in the same row are considered spatial neighbors (left/right adjacency)
  bin_layout = "1-2-3-4-5-6-101-102-7-8-9-10-11-12-13-14-15-16-17-18-103-104-19-20-21-22-23-24-25-26-27-28-29-30-105"
)

```

## 2. Introduction to Synchronicity Analysis

Synchronicity in animal behavior refers to animals performing the same activity at the same time. Understanding synchronicity helps researchers identify social bonds and competition patterns. We will analyze two types of synchronicity: pair-wise co-occurrence and spatial neighbor proximity.

1. **Pair-wise co-occurrence**: How often do 2 animals feed or drink simultaneously, regardless of location (they can be right next to each other or be physically far apart)?
2. **Spatial neighbor proximity**: How often do 2 animals feed or drink right next to each other (spatially close)?

## 3. Prerequisites

This tutorial assumes you have completed data cleaning from [**Tutorial 1: Data Cleaning**](data_cleaning.html). You need cleaned feeding or drinking data.

## 4. Data Preparation

```{r data-prep}
# Load cleaned example data
data(clean_feed)   # For feed-only pair-wise analysis
data(clean_water)  # For water-only pair-wise analysis
data(clean_comb)   # Combined feed + water data for neighbor analysis

# If using your own data from previous tutorials:
# clean_feed <- your_cleaned_feed_data
# clean_water <- your_cleaned_water_data
# clean_comb <- combine_feed_water(clean_feed, clean_water)

```

## 5. Creating Time-Based Activity Matrices

Before analyzing synchronicity, we need to transform visit-based data into time-based matrices.

### Process Feed Data Matrices (for pair-wise analysis)

```{r process-feed-matrices}
# Process feed data to create time-based matrices
# Resolution "sec" means one row per second (more detailed but will generate larger files)
# Resolution "min" means one row per minute (less detailed but will generate smaller files)
feed_matrices <- matrix_process(
  data_list = clean_feed,
  type = "feed",
  resolution = "sec",  # Try "min" for faster processing with less detail
  id_col = id_col2(),
  start_col = start_col2(),
  end_col = end_col2(),
  bin_col = bin_col2(),
  start_weight_col = start_weight_col2(),
  end_weight_col = end_weight_col2(),
  bins_feed = bins_feed2()
)
```

The `matrix_process()` function creates matrices where:

- **Rows** represent time points (by second or minute)
- **Columns** represent individual animals
- **Values** indicate activity status at each time point

`matrix_process()` creates three types of matrices:

1. **Animal activity matrix** (`synch_master_animal2`): Shows which animals are feeding or drinking (i.e., active) at each time point (1 = active, 0 = inactive)
2. **Bin occupancy matrix** (`synch_master_bin2`): Shows which bin each animal is using at each time point (0 = not using any bin)
3. **Feed weight matrix** (`synch_master_feed2`): For feed data only, shows feed weight at each bin over time

The matrices are filtered to include only time points when at least one animal is feeding or drinking, to reduce output data file size.

### Process Water Data Matrices (for pair-wise analysis)

```{r process-water-matrices}
# Process water data similarly
water_matrices <- matrix_process(
  data_list = clean_water,
  type = "drink",
  resolution = "sec",
  id_col = id_col2(),
  start_col = start_col2(),
  end_col = end_col2(),
  bin_col = bin_col2(),
  bins_wat = bins_wat2()
)

```

### Process Combined Feed and Water Data (for neighbor analysis)

For spatial neighbor analysis, we need to use combined feed and water data because water bins are often located between feed bins. Analyzing them separately would miss important spatial relationships.

```{r process-combined-matrices}
# Process combined feed and water data to create time-based matrices
# The clean_comb dataset already contains both feed and water visits
combined_matrices <- matrix_process(
  data_list = clean_comb,
  type = "feed_and_drink",  # Use combined type
  resolution = "sec",
  id_col = id_col2(),
  start_col = start_col2(),
  end_col = end_col2(),
  bin_col = bin_col2(),
  bins_feed = bins_feed2(),
  bins_wat = bins_wat2()
)
```

## 6. Pair-wise Co-Occurrence Analysis

Pair-wise analysis examines how much time every 2 animals spend feeding or drinking simultaneously, regardless of their spatial location.

### Analyze Feed Synchronicity

```{r pair-feed-analysis}
# Analyze pair-wise co-occurrence for feeding
pair_feed_results <- synch_pair_analysis(
  matrix_data = feed_matrices,
  type = "feed",
  resolution = "sec",
  id_col = id_col2()
)

# The results are a list with three matrices:
names(pair_feed_results)

# Look at bout counts for the first day
cat("\nBout matrix (first 5 animals, first day):\n")
bout_matrix_day1 <- pair_feed_results$bout[[1]]
print(bout_matrix_day1[1:5, 1:5])

# Look at total time together
cat("\nTotal time together matrix (seconds, first 5 animals, first day):\n")
time_matrix_day1 <- pair_feed_results$total_time[[1]]
print(time_matrix_day1[1:5, 1:5])

# Look at average duration per bout
cat("\nAverage duration per bout (seconds, first 5 animals, first day):\n")
avg_matrix_day1 <- pair_feed_results$avg_duration[[1]]
print(avg_matrix_day1[1:5, 1:5])
```

### Interpreting Pair-wise Co-Occurrence Results

The results contain three matrices for each day:

- **bout**: Number of distinct feeding or drinking bouts each pair spent together. A new bout starts when animals stop feeding or drinking together for more than 1 second (or 1 minute if using "min" resolution).
- **total_time**: Total time (in seconds or minutes) each pair spent feeding or drinking simultaneously.
- **avg_duration**: Average duration per bout for each pair.

**Note**: Only the upper triangle of each matrix is filled (values above the diagonal) because the matrix is symmetrical.

### Finding Most Synchronized Pairs

```{r find-synchronized-pairs}
# Convert matrices to pair-level data frame for easier analysis
# Use synch_pairs_to_df() to extract upper triangle and sort by total_time
pair_df <- synch_pairs_to_df(
  pair_feed_results,
  min_time = 0,           # Include all pairs with any activity
  sort_by = "total_time", # Sort by time spent together
  decreasing = TRUE       # Most synchronized first
)

cat("Top 10 most synchronized feeding pairs (first day):\n")
print(head(pair_df, 10))
```

Animals that consistently spend high amounts of time feeding or drinking together may have social bonds. On the other hand, low synchronicity despite being in the same herd might indicate avoidance:

```{r avoidance}
# Get least synchronized pairs (sort ascending)
pair_df_least <- synch_pairs_to_df(
  pair_feed_results,
  min_time = 0,
  sort_by = "total_time",
  decreasing = FALSE      # Least synchronized first
)

cat("Top 10 least synchronized feeding pairs (first day):\n")
print(head(pair_df_least, 10))
```

### Analyze Water Drinking Synchronicity

```{r pair-water-analysis}
# Analyze pair-wise co-occurrence for drinking
pair_water_results <- synch_pair_analysis(
  matrix_data = water_matrices,
  type = "drink",
  resolution = "sec",
  id_col = id_col2()
)
```

### Analyze Combined Feed and Water Synchronicity

We can also analyze pair-wise synchronicity using combined feed and water data, which shows how often 2 animals feed or drink at the same time, regardless of their spatial location.

```{r pair-combined-analysis}
# Analyze pair-wise co-occurrence for both feeding and drinking
pair_combined_results <- synch_pair_analysis(
  matrix_data = combined_matrices,
  type = "feed_and_drink",
  resolution = "sec",
  id_col = id_col2()
)

# Extract matrices for the first day (for comparison later)
cat("\nCombined bout matrix (first 5 animals, first day):\n")
combined_bout_matrix <- pair_combined_results$bout[[1]]
print(combined_bout_matrix[1:5, 1:5])

cat("\nCombined total time together (seconds, first 5 animals, first day):\n")
combined_time_matrix <- pair_combined_results$total_time[[1]]
print(combined_time_matrix[1:5, 1:5])
```

## 7. Spatial Neighbor Proximity Analysis

Simply feeding or drinking at the same time doesn't necessarily mean animals have social bonds, especially if they are very far apart at the opposite side of the barn. Neighbor analysis examines how often 2 animals spend at bins right next to each other.

**Important**: Water bins are often located between feed bins, so most often we need to analyze neighbor synchronicity using combined feed and water data to capture true spatial relationships.

### Understanding Bin Layout

The bin layout string defines the spatial arrangement of bins:

- Rows are separated by `\n` (newline)
- Bins within a row are separated by `-`
- Only bins in the same row are considered neighbors (left/right adjacency)

Example: `"1-2-3-4-5-6-101-102-7-8-9-10"` means bins 1, 2, 3, 4, 5, 6, 101, 102, 7, 8, 9, 10 are in a single row. Here, bins 1-30 are feed bins, and bins 101-105 are water bins (with offset of 100).

- Bin 1 is neighbor to bin 2 only
- Bin 101 (water) is neighbor to bins 6 and 102 (water)

```{r show-bin-layout}
cat("Current bin layout:\n")
cat(bin_layout2(), "\n")
```

### Analyze Neighbor Synchronicity (Combined Feed and Water)

```{r neighbor-analysis}
# Analyze spatial neighbor patterns using combined feed and water data
neighbor_results <- synch_neighbor_analysis(
  matrix_data = combined_matrices,
  bin_layout = bin_layout2(),
  type = "feed_and_drink",
  resolution = "sec",
  id_col = id_col2()
)

# The results are a list with three matrices:
names(neighbor_results)

# Look at bout counts for neighbors (first day)
cat("\nNeighbor bout matrix (first 5 animals, first day):\n")
neighbor_bout_matrix <- neighbor_results$bout[[1]]
print(neighbor_bout_matrix[1:5, 1:5])

# Look at total time as neighbors
cat("\nTotal time as neighbors (seconds, first 5 animals, first day):\n")
neighbor_time_matrix <- neighbor_results$total_time[[1]]
print(neighbor_time_matrix[1:5, 1:5])
```

### Comparing Co-Occurrence vs. Neighbor Patterns

```{r compare-patterns}
# Compare the same pair in both analyses (using combined feed+water data)
example_animal1 <- rownames(combined_time_matrix)[1]
example_animal2 <- rownames(combined_time_matrix)[2]

cat(sprintf("Comparison for animals %s and %s (first day):\n
Co-occurrence (feeding OR drinking at same time, any bins): %d bouts, %d seconds total, %.1f seconds per bout on average
Neighbor proximity (feeding OR drinking at adjacent bins): %d bouts, %d seconds total, %.1f seconds per bout on average\n",
  example_animal1, example_animal2,
  combined_bout_matrix[example_animal1, example_animal2],
  combined_time_matrix[example_animal1, example_animal2],
  pair_combined_results$avg_duration[[1]][example_animal1, example_animal2],
  neighbor_bout_matrix[example_animal1, example_animal2],
  neighbor_time_matrix[example_animal1, example_animal2],
  neighbor_results$avg_duration[[1]][example_animal1, example_animal2]
))
```

### Finding Pairs That Prefer Neighboring Bins

```{r neighbor-preference}
# Compare neighbor time to total co-occurrence time (both feed and water)
# This calculates the proportion of total co-occurrence time spent as neighbors
neighbor_df <- synch_neighbor_compare(
  pair_results = pair_combined_results,
  neighbor_results = neighbor_results,
  min_cooccurrence = 0,        # Include all pairs with any co-occurrence
  sort_by = "neighbor_ratio",  # Sort by neighbor preference
  decreasing = TRUE            # Highest preference first
)

cat("Top 10 pairs with highest neighbor preference (first day):\n(Higher ratio = more time at neighboring bins when feeding or drinking together)\n")
print(head(neighbor_df, 10))
```

### Visualize Time Spent Together as a Heatmap

```{r viz-sync-heatmap, fig.width=8, fig.height=7}
# Create a symmetric dataset for the heatmap to show all pairs
# pair_df only contains the upper triangle, so we mirror it
heatmap_data <- rbind(
  pair_df,
  pair_df |> mutate(temp = animal1, animal1 = animal2, animal2 = temp) |> select(-temp)
)

# Convert to symmetric matrix for clustering
all_animals <- unique(c(heatmap_data$animal1, heatmap_data$animal2))
time_matrix <- matrix(0, nrow = length(all_animals), ncol = length(all_animals),
                      dimnames = list(all_animals, all_animals))

# Fill the matrix
for (i in seq_len(nrow(heatmap_data))) {
  time_matrix[as.character(heatmap_data$animal1[i]), 
              as.character(heatmap_data$animal2[i])] <- heatmap_data$total_time[i]
}

# Convert similarity to distance for clustering (higher similarity = lower distance)
# Use max value + 1 as baseline to invert the relationship
max_val <- max(time_matrix, na.rm = TRUE)
dist_matrix <- max_val + 1 - time_matrix
diag(dist_matrix) <- 0  # Set diagonal to 0

# Perform hierarchical clustering
dist_obj <- as.dist(dist_matrix)
hc <- hclust(dist_obj, method = "ward.D2")
animal_order <- hc$labels[hc$order]

# Build a complete grid (including diagonal and zero-time pairs) for plotting
plot_grid <- expand.grid(
  animal1 = animal_order,
  animal2 = animal_order,
  stringsAsFactors = FALSE
)
plot_grid$total_time <- time_matrix[cbind(plot_grid$animal1, plot_grid$animal2)]
plot_grid$animal1 <- factor(plot_grid$animal1, levels = animal_order)
plot_grid$animal2 <- factor(plot_grid$animal2, levels = animal_order)

# Visualize the total time spent together as a heatmap using viridis palette
ggplot(plot_grid, aes(x = animal1, y = animal2, fill = total_time)) +
  geom_tile() +
  scale_fill_viridis_c(option = "viridis", name = "Total Time\n(seconds)", direction = -1) +
  labs(
    title = "Pair-wise Synchronicity Heatmap",
    subtitle = "Total time spent feeding together (clustered by similarity, first day)",
    x = "Animal ID",
    y = "Animal ID"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )
```

### Visualize Time Spent Feeding as Neighbors as a Heatmap

```{r viz-neighbor-heatmap, fig.width=8, fig.height=7}
# Convert neighbor results to data frame
neighbor_df_viz <- synch_pairs_to_df(
  synch_results = neighbor_results,
  min_time = 0,
  sort_by = "total_time",
  decreasing = TRUE
)

# Create a symmetric dataset for the heatmap to show all pairs
neighbor_heatmap_data <- rbind(
  neighbor_df_viz,
  neighbor_df_viz |> mutate(temp = animal1, animal1 = animal2, animal2 = temp) |> select(-temp)
)

# Convert to symmetric matrix for clustering
all_animals_neighbor <- unique(c(neighbor_heatmap_data$animal1, neighbor_heatmap_data$animal2))
neighbor_time_matrix <- matrix(0, nrow = length(all_animals_neighbor), 
                                ncol = length(all_animals_neighbor),
                                dimnames = list(all_animals_neighbor, all_animals_neighbor))

# Fill the matrix
for (i in seq_len(nrow(neighbor_heatmap_data))) {
  neighbor_time_matrix[as.character(neighbor_heatmap_data$animal1[i]), 
                       as.character(neighbor_heatmap_data$animal2[i])] <- 
    neighbor_heatmap_data$total_time[i]
}

# Convert similarity to distance for clustering
max_val_neighbor <- max(neighbor_time_matrix, na.rm = TRUE)
neighbor_dist_matrix <- max_val_neighbor + 1 - neighbor_time_matrix
diag(neighbor_dist_matrix) <- 0

# Perform hierarchical clustering
neighbor_dist_obj <- as.dist(neighbor_dist_matrix)
neighbor_hc <- hclust(neighbor_dist_obj, method = "ward.D2")
neighbor_animal_order <- neighbor_hc$labels[neighbor_hc$order]

# Build a complete grid (including diagonal and zero-time pairs) for plotting
neighbor_plot_grid <- expand.grid(
  animal1 = neighbor_animal_order,
  animal2 = neighbor_animal_order,
  stringsAsFactors = FALSE
)
neighbor_plot_grid$total_time <- neighbor_time_matrix[cbind(neighbor_plot_grid$animal1, neighbor_plot_grid$animal2)]
neighbor_plot_grid$animal1 <- factor(neighbor_plot_grid$animal1, levels = neighbor_animal_order)
neighbor_plot_grid$animal2 <- factor(neighbor_plot_grid$animal2, levels = neighbor_animal_order)

# Visualize the total time spent feeding as neighbors as a heatmap
ggplot(neighbor_plot_grid, aes(x = animal1, y = animal2, fill = total_time)) +
  geom_tile() +
  scale_fill_viridis_c(option = "viridis", name = "Total Time\n(seconds)", direction = -1) +
  labs(
    title = "Neighbor Synchronicity Heatmap",
    subtitle = "Total time spent feeding or drinking as neighbors (clustered by similarity, first day)",
    x = "Animal ID",
    y = "Animal ID"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )
```

## 8. Code Cheatsheet

```{r cheatsheet, eval=FALSE}
#' Copy and modify these code blocks for your own analysis!

# ---- SETUP: Global Variables (REQUIRED FIRST!) ----
library(moo4feed)
library(ggplot2)
library(dplyr)

# Set up column names and bin configuration to match your data structure
set_global_cols(
  # Time zone
  tz = "America/Vancouver",        # Your timezone
  
  # Column names in your cleaned data
  id_col = "cow",
  trans_col = "transponder",
  start_col = "start",
  end_col = "end",
  bin_col = "bin",
  dur_col = "duration",
  intake_col = "intake",
  start_weight_col = "start_weight",
  end_weight_col = "end_weight",
  
  # Bin settings
  bins_feed = 1:30,                # All feed bin IDs in your barn
  bins_wat = 1:5,                  # All water bin IDs in your barn
  bin_offset = 100,                # Add this to water bin IDs to avoid conflicts
  
  # Physical bin layout for spatial neighbor analysis
  # Rows separated by "\n", bins within rows separated by "-"
  # Only bins in the same row are considered spatial neighbors (left/right adjacency)
  bin_layout = "1-2-3-4-5-6-101-102-7-8-9-10-11-12-13-14-15-16-17-18-103-104-19-20-21-22-23-24-25-26-27-28-29-30-105"
)

# ---- STEP 1: Load Cleaned Data ----
# Use data from Tutorial 1: Data Cleaning
# For demo data:
data(clean_feed)   # For feed-only pair-wise analysis
data(clean_water)  # For water-only pair-wise analysis
data(clean_comb)   # Combined feed + water data for neighbor analysis

# For your own data:
# clean_feed <- your_cleaned_feed_data    # From data cleaning tutorial
# clean_water <- your_cleaned_water_data  # From data cleaning tutorial
# clean_comb <- combine_feed_water(clean_feed, clean_water)  # Combine them

# ---- STEP 2: Create Time-Based Activity Matrices ----
# Process feed data to create time-based matrices (for pair-wise analysis)
feed_matrices <- matrix_process(
  data_list = clean_feed,          # Your cleaned feed data
  type = "feed",                   # Data type: "feed" or "drink"
  resolution = "sec",              # Time resolution: "sec" (detailed) or "min" (faster)
  id_col = id_col2(),              # Animal ID column (from global vars)
  start_col = start_col2(),        # Visit start time column (from global vars)
  end_col = end_col2(),            # Visit end time column (from global vars)
  bin_col = bin_col2(),            # Bin ID column (from global vars)
  start_weight_col = start_weight_col2(),  # Start weight column (from global vars)
  end_weight_col = end_weight_col2(),      # End weight column (from global vars)
  bins_feed = bins_feed2()         # Valid feed bin IDs (from global vars)
)

# Process water data to create time-based matrices (for pair-wise analysis)
water_matrices <- matrix_process(
  data_list = clean_water,         # Your cleaned water data
  type = "drink",                  # Data type for water/drinking
  resolution = "sec",              # Must match resolution used for feed_matrices
  id_col = id_col2(),
  start_col = start_col2(),
  end_col = end_col2(),
  bin_col = bin_col2(),
  bins_wat = bins_wat2()           # Valid water bin IDs (from global vars)
)

# IMPORTANT: Process combined feed and water data for neighbor analysis
# Water bins are often between feed bins, so analyze them together for spatial relationships
# Use clean_comb dataset (already combined) or create your own with combine_feed_water()
combined_matrices <- matrix_process(
  data_list = clean_comb,          # Combined feed and water data
  type = "feed_and_drink",         # Combined type
  resolution = "sec",              # Must match resolution used above
  id_col = id_col2(),
  start_col = start_col2(),
  end_col = end_col2(),
  bin_col = bin_col2(),
  bins_feed = bins_feed2(),
  bins_wat = bins_wat2()
)

# ---- STEP 3: Pair-wise Co-Occurrence Analysis ----
# Analyze feeding synchronicity (animals feeding at same time, any bins)
pair_feed_results <- synch_pair_analysis(
  matrix_data = feed_matrices,     # Output from matrix_process()
  type = "feed",                   # Data type: "feed" or "drink"
  resolution = "sec",              # Must match matrix_process() resolution
  id_col = id_col2()               # Animal ID column (from global vars)
)

# Results contain three matrices/lists:
names(pair_feed_results)  # bout, total_time, avg_duration

# Analyze drinking synchronicity (animals drinking at same time, any bins)
pair_water_results <- synch_pair_analysis(
  matrix_data = water_matrices,    # Output from matrix_process()
  type = "drink",                  # Data type for water/drinking
  resolution = "sec",              # Must match matrix_process() resolution
  id_col = id_col2()
)

# IMPORTANT: Analyze combined feed+water synchronicity for comparison with neighbor analysis
pair_combined_results <- synch_pair_analysis(
  matrix_data = combined_matrices, # Output from matrix_process() with combined data
  type = "feed_and_drink",         # Combined type
  resolution = "sec",              # Must match matrix_process() resolution
  id_col = id_col2()
)

# Example: inspect first day matrices from combined pair analysis
combined_bout_matrix <- pair_combined_results$bout[[1]]
combined_time_matrix <- pair_combined_results$total_time[[1]]

# Access pair analysis results (matrices or lists of matrices for multi-day)
# For single day: each element is a matrix
# For multi-day: each element is a list with one matrix per day
bout_counts <- pair_feed_results$bout           # Number of distinct bouts together
total_times <- pair_feed_results$total_time    # Total time together (seconds or minutes)
avg_durations <- pair_feed_results$avg_duration # Average duration per bout

# Example: Access first day's bout matrix
if (is.list(pair_feed_results$bout)) {
  day1_bouts <- pair_feed_results$bout[[1]]    # Multi-day data
} else {
  day1_bouts <- pair_feed_results$bout         # Single-day data
}

# ---- STEP 4: Spatial Neighbor Proximity Analysis ----
# Check bin layout
cat("Current bin layout:\n")
cat(bin_layout2(), "\n")

# IMPORTANT: Use combined matrices for neighbor analysis!
# Water bins are often between feed bins, so we analyze them together
neighbor_results <- synch_neighbor_analysis(
  matrix_data = combined_matrices, # Output from matrix_process() with combined data
  bin_layout = bin_layout2(),      # Physical bin arrangement (from global vars)
  type = "feed_and_drink",         # Combined type
  resolution = "sec",              # Must match matrix_process() resolution
  id_col = id_col2()               # Animal ID column (from global vars)
)

# Results contain three matrices/lists:
names(neighbor_results)  # bout, total_time, avg_duration

# Access neighbor analysis results
neighbor_bouts <- neighbor_results$bout         # Bouts at neighboring bins
neighbor_times <- neighbor_results$total_time  # Time at neighboring bins
neighbor_avg <- neighbor_results$avg_duration  # Average neighbor bout duration

# Compare co-occurrence vs neighbor patterns (example)
# Extract matrices for the first day
combined_bout_matrix <- pair_combined_results$bout[[1]]
combined_time_matrix <- pair_combined_results$total_time[[1]]
neighbor_bout_matrix <- neighbor_results$bout[[1]]
neighbor_time_matrix <- neighbor_results$total_time[[1]]

example_animal1 <- rownames(combined_time_matrix)[1]
example_animal2 <- rownames(combined_time_matrix)[2]
cat(sprintf("Comparison for animals %s and %s (first day):\n
Co-occurrence (feeding OR drinking at same time, any bins): %d bouts, %d seconds total, %.1f seconds per bout on average
Neighbor proximity (feeding OR drinking at adjacent bins): %d bouts, %d seconds total, %.1f seconds per bout on average\n",
  example_animal1, example_animal2,
  combined_bout_matrix[example_animal1, example_animal2],
  combined_time_matrix[example_animal1, example_animal2],
  pair_combined_results$avg_duration[[1]][example_animal1, example_animal2],
  neighbor_bout_matrix[example_animal1, example_animal2],
  neighbor_time_matrix[example_animal1, example_animal2],
  neighbor_results$avg_duration[[1]][example_animal1, example_animal2]
))

# ---- STEP 5: Convert Matrices to Tidy Data Frames ----
# Convert pair analysis results to data frame for easier analysis
pair_df <- synch_pairs_to_df(
  synch_results = pair_feed_results,  # Output from synch_pair_analysis()
  min_time = 0,                       # Minimum time threshold (0 = include all)
  sort_by = "total_time",             # Column to sort by: "total_time", "bouts", "avg_duration"
  decreasing = TRUE                   # TRUE = descending, FALSE = ascending
)

# Convert neighbor analysis results to data frame
neighbor_df <- synch_pairs_to_df(
  synch_results = neighbor_results,  # Output from synch_neighbor_analysis()
  min_time = 0,                      # Minimum neighbor time threshold
  sort_by = "total_time",            # Sort by time at neighboring bins
  decreasing = TRUE
)

# ---- STEP 6: Find Most/Least Synchronized Pairs ----
# Get top 10 most synchronized feeding pairs
top_pairs <- head(pair_df, 10)
print(top_pairs)

# Get top 10 least synchronized feeding pairs
least_pairs <- synch_pairs_to_df(
  pair_feed_results,
  min_time = 0,
  sort_by = "total_time",
  decreasing = FALSE               # Least synchronized first
)
print(head(least_pairs, 10))

# Filter pairs with high synchronicity (e.g., >100 seconds together)
high_sync_pairs <- pair_df[pair_df$total_time > 100, ]

# ---- STEP 7: Compare Neighbor Preference to Total Co-Occurrence ----
# IMPORTANT: Compare combined feed+water co-occurrence with neighbor proximity
# This ensures we're comparing the same activity types (both feed and water)
neighbor_compare <- synch_neighbor_compare(
  pair_results = pair_combined_results, # Combined co-occurrence results
  neighbor_results = neighbor_results,  # Neighbor results (combined feed+water)
  min_cooccurrence = 0,                # Minimum co-occurrence time threshold (0 = all pairs)
  sort_by = "neighbor_ratio",          # Sort by neighbor preference ratio
  decreasing = TRUE                    # Highest preference first
)

# neighbor_compare contains:
# - cooccurrence_time: Total time active together (feed OR water) anywhere
# - neighbor_time: Time at neighboring bins (feed OR water)
# - neighbor_ratio: neighbor_time / cooccurrence_time (0 to 1)

# Find pairs that prefer neighboring bins (high ratio)
high_neighbor_preference <- neighbor_compare[neighbor_compare$neighbor_ratio > 0.5, ]

# ---- STEP 8: Visualize Results ----
# Create a symmetric dataset for the heatmap to show all pairs
heatmap_data <- rbind(
  pair_df,
  pair_df |> mutate(temp = animal1, animal1 = animal2, animal2 = temp) |> select(-temp)
)

# Convert to symmetric matrix for clustering
all_animals <- unique(c(heatmap_data$animal1, heatmap_data$animal2))
time_matrix <- matrix(0, nrow = length(all_animals), ncol = length(all_animals),
                      dimnames = list(all_animals, all_animals))

# Fill the matrix
for (i in seq_len(nrow(heatmap_data))) {
  time_matrix[as.character(heatmap_data$animal1[i]), 
              as.character(heatmap_data$animal2[i])] <- heatmap_data$total_time[i]
}

# Convert similarity to distance for clustering (higher similarity = lower distance)
max_val <- max(time_matrix, na.rm = TRUE)
dist_matrix <- max_val + 1 - time_matrix
diag(dist_matrix) <- 0  # Set diagonal to 0

# Perform hierarchical clustering
dist_obj <- as.dist(dist_matrix)
hc <- hclust(dist_obj, method = "ward.D2")
animal_order <- hc$labels[hc$order]

# Build a complete grid (including diagonal and zero-time pairs) for plotting
plot_grid <- expand.grid(
  animal1 = animal_order,
  animal2 = animal_order,
  stringsAsFactors = FALSE
)
plot_grid$total_time <- time_matrix[cbind(plot_grid$animal1, plot_grid$animal2)]
plot_grid$animal1 <- factor(plot_grid$animal1, levels = animal_order)
plot_grid$animal2 <- factor(plot_grid$animal2, levels = animal_order)

# Visualize the total time spent together as a heatmap using viridis palette
ggplot(plot_grid, aes(x = animal1, y = animal2, fill = total_time)) +
  geom_tile() +
  scale_fill_viridis_c(option = "viridis", name = "Total Time\n(seconds)", direction = -1) +
  labs(
    title = "Pair-wise Synchronicity Heatmap",
    subtitle = "Total time spent feeding together (clustered by similarity, first day)",
    x = "Animal ID",
    y = "Animal ID"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )

# Visualize neighbor feeding time as a heatmap (clustered by similarity)
# Convert neighbor results to data frame
neighbor_df_viz <- synch_pairs_to_df(
  synch_results = neighbor_results,
  min_time = 0,
  sort_by = "total_time",
  decreasing = TRUE
)

# Create a symmetric dataset for the neighbor heatmap
neighbor_heatmap_data <- rbind(
  neighbor_df_viz,
  neighbor_df_viz |> dplyr::mutate(temp = animal1, animal1 = animal2, animal2 = temp) |> dplyr::select(-temp)
)

# Convert to symmetric matrix for clustering
all_animals_neighbor <- unique(c(neighbor_heatmap_data$animal1, neighbor_heatmap_data$animal2))
neighbor_time_matrix <- matrix(0, nrow = length(all_animals_neighbor), 
                                ncol = length(all_animals_neighbor),
                                dimnames = list(all_animals_neighbor, all_animals_neighbor))

# Fill the matrix
for (i in seq_len(nrow(neighbor_heatmap_data))) {
  neighbor_time_matrix[as.character(neighbor_heatmap_data$animal1[i]), 
                       as.character(neighbor_heatmap_data$animal2[i])] <- 
    neighbor_heatmap_data$total_time[i]
}

# Convert similarity to distance for clustering
max_val_neighbor <- max(neighbor_time_matrix, na.rm = TRUE)
neighbor_dist_matrix <- max_val_neighbor + 1 - neighbor_time_matrix
diag(neighbor_dist_matrix) <- 0

# Perform hierarchical clustering
neighbor_dist_obj <- as.dist(neighbor_dist_matrix)
neighbor_hc <- hclust(neighbor_dist_obj, method = "ward.D2")
neighbor_animal_order <- neighbor_hc$labels[neighbor_hc$order]

# Build a complete grid (including diagonal and zero-time pairs) for plotting
neighbor_plot_grid <- expand.grid(
  animal1 = neighbor_animal_order,
  animal2 = neighbor_animal_order,
  stringsAsFactors = FALSE
)
neighbor_plot_grid$total_time <- neighbor_time_matrix[cbind(neighbor_plot_grid$animal1, neighbor_plot_grid$animal2)]
neighbor_plot_grid$animal1 <- factor(neighbor_plot_grid$animal1, levels = neighbor_animal_order)
neighbor_plot_grid$animal2 <- factor(neighbor_plot_grid$animal2, levels = neighbor_animal_order)

# Visualize the total time spent feeding as neighbors as a heatmap
ggplot(neighbor_plot_grid, aes(x = animal1, y = animal2, fill = total_time)) +
  geom_tile() +
  scale_fill_viridis_c(option = "viridis", name = "Total Time\n(seconds)", direction = -1) +
  labs(
    title = "Neighbor Synchronicity Heatmap",
    subtitle = "Total time spent feeding or drinking as neighbors (clustered by similarity, first day)",
    x = "Animal ID",
    y = "Animal ID"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )

```

---
